Peer reviewers altered their recommendation based on whether they were cited or wanted to be cited. A matched study of open peer review at four journals

Version 1

Abstract

Abstract

Peer reviewers judge the validity and quality of new research. These judgements would ideally be impartial, but some reviewers may give a more favourable review if they are cited in the article because the authors have recognised their work and because citations are a valuable academic currency. Reviewers sometimes request self-citations to their own work, which may be justified as reviewers should be relevant experts. However, some self-citation requests may be unethical, with reviewers exploiting the authors’ need to publish. We examined whether citations to a reviewer and self-citations influenced their peer review. We used a matched design at four journals that use open peer review and make all article versions available. Our sample included more than 37,000 peer reviews, with 13% where the reviewer was cited in the article and 6% where the reviewer included a self-citation to their work in their review. Reviewers who were cited were more likely to approve the article, with an odds ratio of 1.61 compared with reviewers who were not cited (adjusted 99.4% CI: 1.16 to 2.23). Reviewers who suggested a self-citation were much less likely to approve the article, with an odds ratio of 0.15 compared with reviewers with no self-citations (adjusted 99.4% CI: 0.08 to 0.30). Reviewers who requested and received a citation were much more likely to approve the article compared to reviewers whose request was disregarded (odds ratio of 3.5, 95% CI: 2.0 to 6.1). Some reviewers’ recommendations are dependent on whether they are cited or want to be cited. Self-citation requests can turn peer review into a transaction rather than an objective critique of the article.

Introduction

In 2024, a published peer reviewed article included this remarkable sentence: “As strongly requested by the reviewers, here we cite some references [35–47] although they are completely irrelevant to the present work” [1]. This was a rare public example of coerced citations, where a reviewer exploits the peer review process to increase their citation counts and hence further their own career [2–4]. Reviewers should be relevant experts, so some suggestions to cite their work will be appropriate. However, excessive requests for self-citations or requests to cite unrelated work are unethical [5–9]. Coerced citations can also come from editors trying to boost their journal’s ranking [10–12].

Coerced citations are reported as a common problem in peer review. In author surveys, two-thirds reported pressure from peer reviewers to cite unrelated articles [13] and 23% had experienced a reviewer that “required them to include unnecessary references to their publication(s)” [14]. Publishers have investigated whether “hundreds of researchers” have manipulated the peer review process to increase their own citations [15]. Some reviewers may be exploiting their power over authors who “have a strong incentive to […] accept all ‘suggestions’ by the referees even if one knows that they are misleading or even incorrect” [16].

As reviewers are often in the same field as the authors, they may already be cited in the article without the need for coerced citations. Reviewers who are cited may give a more favourable peer review and be more willing to overlook flaws [17, 18]. Some authors may try to exploit this using “referee baiting” [3] or “flattery citations” [19] by favourably citing a reviewer’s work.

The interactions during peer review between authors and reviewers can determine whether an article is accepted [20] and what results are included in the published version [21]. Given the importance of peer review for science, studies that examine how peer review works in practice are needed [22–26]. Here, we examine interactions between peer reviewers and authors using four journals that publish all article versions and all peer reviews. We had two research questions:

Do peer reviewers give a more or less favourable recommendation when they are cited in the article?

Do peer reviewers give a more or less favourable recommendation when their review includes a self-citation?

Methods

Journal selection

We studied journals from the publisher F1000 as their journals use open peer review with signed reviewers. F1000 journals use a publish–review–curate model [27], meaning all versions of the article are publicly available, including versions updated after peer review. This allowed us to examine the interactions between authors and reviewers throughout the peer review process. We selected four F1000 journals that each had over 100 articles. Some characteristics of the four journals are given in Table 1. Three journals were created to support funders.

The peer review process used by F1000 journals differs from most standard journals. The journals do not use academic editors, but do have in-house editors who manage articles but do not make editorial decisions. This means that most interactions during peer review are between authors and reviewers directly. In-house editors perform checks prior to the first version of the article being published and at F1000Research this results in 40 to 50% submissions being rejected (personal communication, F1000 staff). Up to mid-2024, the authors were asked to identify potential reviewers who were qualified experts with no competing interests [28]. Since mid-2024, reviewer identification is made in-house, although authors can suggest reviewers.

Reviewers are asked to recommend one of three categories: Approved, Approved with reservations, and Not approved. For brevity, we refer to ‘Approved with Reservations’ as ‘Reservations’. An article is indexed once it receives two ‘Approved’ or two ‘Reservations’ and one ‘Approved’. The guidelines for recommending Approved are: “the aims and research methods are adequate; results are presented accurately, and the conclusions are justified and supported by the presented data” [29]. Peer reviewers are asked to assess the validity of an article’s content, rather than novelty or interest levels, an approach designed to combat publication bias [30].

All four journals have a peer reviewer code of conduct and state that reviewers should familiarise themselves with the ethical guidelines for peer reviewers by the Committee On Publication Ethics [31]. The journals’ guidelines for reviewers include the following: “reviewers should explicitly state their reasoning when asking authors to cite their own work”.

Data extraction

We extracted data on authors and articles from the OpenAlex database (https://openalex.org/) and directly from the four journals. OpenAlex combines scholarly data from multiple sources, including ORCID – a unique identifier for researchers, Microsoft Academic, Crossref and PubMed. A recent study compared OpenAlex with the two commonly used proprietary bibliometric databases of Web of Science and Scopus for the years 2015 to 2022 [32]. The results were mixed, but OpenAlex had better ORCID coverage and covered more Digital Object Identifiers (DOIs) – the unique identifier for publications. We accessed OpenAlex using the openalexR package [33, 34]. We used each journal’s application programming interface (API) to extract data on the articles and peer reviews. The data were extracted in four stages:

Searches were made using the APIs of the four journals to find all articles published between 1 Jan 2012 and 28 May 2025, with the start date to capture all potential articles.

For each article, the following data were downloaded in XML format:

The article’s publication date and version number

The reviewers’ names and ORCIDs (if available)

The text of all reviews and the reviewers’ recommendations

The DOIs and PMIDs (PubMed IDs) from the article’s reference list

The DOIs and PMIDs of any articles cited by the reviewers. The online peer review system at F1000 journals includes the DOI of any article cited in the review, which facilitates the identification of self-citations.

Articles were excluded if:

They were not peer reviewed or had yet to receive any reviews

The reference list was empty

The reviewers’ publication histories were collected from OpenAlex using their name, institution and ORCID (if available). Reviews were excluded if there was no record for the reviewer in OpenAlex, or if the reviewer had no published articles as there was no potential for them to be cited or request a self-citation.

Study design

We used two predictor variables about the reviewer:

The number of times they were cited in the article (0,1, 2,…).

The number of times they included self-citations to their own work in their review (0,1, 2,…).

We fitted both predictors as linear, but reviewers may behave differently with any citation rather than a linear change, and hence we also fitted both predictors as a binary “none versus any” (0 vs 1, 2, …). We compared the linear and binary alternatives using the Akaike Information Criterion (AIC) to find the parameterisation that best fitted the data [35].

We matched on article and version to control for confounding by any characteristics of the article [36]; for example, the article’s topic or writing style. Hence, we compared two or more independent reviewers who considered the same article.

All analyses were stratified by article version, using the first version only or the second and subsequent versions. This is because the reviewers are unknown to the authors for the first version, but from the second version onwards, the authors will know the reviewers as the journals use signed peer reviews. This knowledge could alter the behaviour of authors and reviewers.

Any confounding by the characteristics of the articles was controlled by the matched design, but confounding by the characteristics of the reviewers remains possible [18]. We considered the potential confounders of the reviewer’s experience and reviewer’s country. More experienced reviewers will likely be cited more often (on average) and could be more or less strict in their recommendations. The reviewer’s country is a potential confounder due to large differences in citation counts by country [37] and potential differences in recommendations by country [38].

Some reviews were performed by reviewers together with co-reviewers, who were usually less experienced. Our primary analysis excluded the co-reviewers, but we included them in a sensitivity analysis where we created combined versions of the two independent variables using the sum of citations to reviewers and co-reviewers, and the sum of self-citations from the reviewers and co-reviewers.

The study design is summarised in Figure 1

Graphical summary of the study design for research question 1 showing a dummy article and two reviews.

In the first version of the article, the reviewer Smith (blue) is cited whilst Jones (purple) is not. For the second version of the article, the authors are now aware that Jones is a reviewer and Jones has been cited. The reviewers’ recommendations are the outcome and are colour-coded as Not approved (red), Reservations (orange) and Approved (green). We tested whether citations to the reviewer in the article influenced their recommendation. The matched design means that only reviewers of the same article are compared (here, Smith and Jones) and the overall effect is estimated by aggregating over multiple matched comparisons. Research question 2 used the same design but examined self-citations in the reviews.

Statistical methods

We used conditional logistic regression to examine the associations between the citations to the reviewer and their ordinal recommendation (Approved → Reservations → Not approved) while matching the article and the version [39]. Conditional logistic regression requires a binary dependent variable; hence, we fitted two related models that examined the odds of:

“Approved” compared with “Reservations” or “Not approved”.

“Approved” or “Reservations” compared with “Not approved”,

These two models tested the same hypothesis, hence we adjusted for multiple testing. We also used repeated testing because of the stratification by article version and the two formulations of the predictors (linear or none versus any). Since we used 8 (2 × 2 × 2) tests, we displayed all the results using 99.4% confidence intervals instead of 95.0% intervals, which is a 5% type I error divided by eight tests.

In an unplanned analysis, we examined the association between the reviewer’s recommendation and whether they included citations to work other than self-citations. This was added to examine whether the citations were used to highlight important errors and/or missing context in the article.

In sensitivity analyses, we controlled for potential confounding due to the reviewer. We used the reviewer’s number of published articles as a proxy for their experience. This association could be non-linear; for example, a diminishing effect for more experienced reviewers, so we examined six fractional polynomials of the reviewers’ number of articles and used the AIC to select the best association [40].

In a second sensitivity analysis, we planned to use a frailty model for the reviewers’ countries [41]. However, this model often failed to converge, potentially because there were many countries and some countries had relatively small numbers of reviewers. Hence, we instead used a leave-one-out analysis for each of the top ten most common countries and determined if the results were noticeably different.

Outliers were not excluded. No data were missing in the analysis data set. We tested for potential bias from reviewers who were excluded because they had no data in OpenAlex using an elastic net with potential predictors of article version, article date, reviewer’s country, and role (reviewer or co-reviewer) and dependent variable of reviewer excluded (yes/no) [42].

Text analysis

We examined how self-citations were justified in the reviews and whether the text differed according to the reviewer’s recommendation. For an initial view of self-citations, we randomly selected 20 reviews that included a self-citation and extracted the most relevant sentence.

To analyse the review text, we first extracted the 100 most commonly used words across all reviews. To standardise the text, all words were transformed into tokens, with stop-words removed and then stemmed. We then tested which of the 100 words were associated with recommending Approved versus Reservations or Not approved amongst those reviewers who included a self-citation. We chose the set of words using an elastic net with 10-fold cross-validation and selected a parsimonious model by using the lambda within one standard error of the minimum cross-validated error [43]. To get uncertainty intervals for the estimates, we fitted a Bayesian model with the set of words selected by the elastic net and using a sceptical Normal prior centred on zero to create shrinkage.

Sample size

We aimed for a sample size of approximately 5,000 articles and assumed that half would be the first version, giving a sample size of 2,500 articles for the analysis using the first version only [44]. In 1,000 simulations, this gave an 89.1% power to detect an odds ratio of 1.5 using conditional logistic regression for a reviewer who recommended a higher category (Approved → Reservations → Not approved) when they were cited. We assumed that 15% of articles would include a citation to the reviewer. Eighty percent of the simulated articles had two reviews, and the remaining 20% had three reviews. Based on preliminary data from two journals, we assumed that the reviewers’ recommendations would have a ratio for Approved:Reservations:Not approved of 70:24:6.

Reproducibility

Research question 1 was pre-registered using As Predicted on 20 May 2024 [44]. Research question 2 was formulated during data collection but before any data analysis and used the same study design and statistical methods as question 1.

All data extraction and analyses were conducted using R version 4.4.1 [45]. The data and R code are available on GitHub [46].

Results

A flow chart of the included reviews is shown in Supplement S.1. The final sample size was over 37,000 reviews. There were more than 2,000 articles that were not included because they had not yet been peer reviewed, especially recent articles. More than 900 reviewers did not have a record in OpenAlex and so could not be included. These missing reviewers were more likely to be from older articles and more likely to be co-reviewers.

Descriptive statistics on the included reviews are in Table 2. The reviewers were cited at least once in 13% of the articles and 6% of the reviews included a self-citation. Most reviews recommended “Approved” (54%), with only 8% recommending “Not approved” which is low compared with many journals; however, 40 to 50% of submissions are rejected before articles are sent for peer review (personal communication, F1000 staff). The reviewers were relatively experienced, with a median number of papers of 55.

The binary predictor for citations of “any versus none” had a generally better fit to the data compared to the linear predictor (Supplement S.2). This indicates that for most reviewers receiving any citation is important, and there is no linear increase for two or more citations. The following results are for the binary predictor “any versus none”, with the results using a linear predictor in Supplement S.3.

Reviewers who were cited were more likely to approve the article, but only after version 1 (Fig 2 and Table 3). If a reviewer was cited in any versions after version 1, the odds ratio for recommending Approved versus Reservations or Not approved was 1.61 (adjusted 99.4% CI 1.16 to 2.23).

Odds ratios and probabilities for reviewers giving a more or less favourable recommendation depending on whether they were cited in the article.

The top row examines Approved vs Reservations or Not approved, and the bottom row examines Approved or Reservations vs Not approved. The figures show the mean (dot) and adjusted 99.4% confidence intervals (horizontal lines). All models were split by article version. The odds ratios and probabilities show the same results but on different scales.

Reviewers who requested a self-citation were much less likely to approve the article for all versions (Fig 3 and Table 3). The odds ratio for recommending Approved versus Reservations or Not approved was 0.57 (99.4% CI 0.44 to 0.73) for version 1 and strengthened to 0.15 (99.4% CI 0.08 to 0.30) for versions 2+. The less favourable recommendation was only for the approval of the article and the odds ratios for Approved or Reservations versus Not approved were much closer to 1.

Odds ratios and probabilities for reviewers giving a more or less favourable recommendation if they included citations to their own research in their review.

The top row examines Approved vs Reservations or Not approved, and the bottom row examines Approved or Reservations vs Not approved. The figures show the mean (dot) and adjusted 99.4% confidence intervals (horizontal lines). All models were split by article version. The odds ratios and probabilities show the same results but on different scales.

In an unplanned analysis, we examined the behaviour of reviewers in the first two versions of the article. We examined the 441 reviews where the reviewer was not cited in version 1 of the article and requested a self-citation in their first review. The reviewers who were then cited in version 2 recommended approval for 92%, compared to 76% for reviewers who were not cited (odds ratio = 3.5, 95% CI: 2.0 to 6.1). This analysis did not use matching.

In an unplanned analysis, we examined whether the reviewers’ recommendations depended on citations to research other than their own by excluding self-citations. Reviewers who included citations in their review were much more likely not to approve the article (Figure 4), which was similar to the association with self-citations (Figure 3). However, reviewers who included citations other than self-citations were also much more likely to recommend “Not approved”, as shown by the lower odds of “Not approved” or “Reservations” versus “Not approved”. This association was not seen using self-citations (Figure 3)

Odds ratios and probabilities for reviewers giving a more or less favourable recommendation depending on if they included citations to other research in their review.

The top row examines Approved vs Reservations or Not approved, and the bottom row examines Approved or Reservations vs Not approved. The figures show the mean (dot) and adjusted 99.4% confidence intervals (horizontal lines). All models were split by article version. The odds ratios and probabilities show the same results but on different scales.

Sensitivity analyses

The odds ratios when including co-reviewers with reviewers were similar to the odds ratios when using reviewers only (Supplement S.4).

We found no evidence that the reviewers’ publication numbers or country confounded the associations between citations and recommendations (Supplement S.5).

Text analyses of reviewers’ comments

A random sample of how reviewers requested self-citations found some vague justifications (Supplement S.8); for example, “Here are some additional publications you might consider referencing”. Other sentences adhered to the publisher’s guidelines for reviewers, as specific reasoning was provided for self-citation [7]. One reviewer thanked the authors for a previous citation. Three reviews did not have a relevant sentence. One reviewer almost certainly used AI to write their review as it included the phrase “Certainly! Here are some potential review questions for the manuscript” [47]; this review included six self-citations with no justifications.

Reviewers who included a self-citation were more likely to use the words “need” and “please” when not approving the article (Figure 5). In contrast, the words “genome” and “might” were the most strongly associated with the reviewers’ approval.

Words in the reviewers’ comments that were associated with approving the article or not for reviewers who included a self-citation (n = 2, 710).

The words were selected using an elastic net that started with the 100 most commonly used review words with 28 retained. The estimates from the elastic net are shown as empty circles and the mean estimates and 95% credible intervals from a Bayesian model as shown as a solid circle and horizontal line. The axis label shows the stemmed word and most common whole word in brackets.

To examine how often open peer reviews were viewed, we took a random sample of 200 reviews from the four journals and found that, on average, they were viewed just 1.2 times per year (Supplement S.7).

Discussion

Our results show strong evidence that some reviewers have a transactional view of peer review, with their final approval dependent on citations to their work. These reviewers are exploiting the pressure on authors to “publish or perish”. Under this pressure, many authors may oblige and add the suggested citations, especially since adding another citation may only require a minor edit to their article [48]. Both sides gain from this transaction, as the authors get an indexed publication and the reviewer gets a citation.

Some reviewers who withhold their approval and request a self-citation may be justified, as they may be highlighting important errors or missing context in the article. Self-citations can be justified when the authors have made a “large scholarly oversight” [8]. However, our matched design shows that other reviewers who evaluated the same article and who did not self-cite were often willing to approve the article, suggesting that the concerns of the self-citing reviewer were specific to them and not serious enough to be noticed by other reviewers.

Reviewers including citations to other research (excluding self-citations) were also less likely to approve the article, supporting the idea that citations (self or otherwise) are highlighting important errors or missing context. However, reviewers citing other research were also much more likely to recommend “Not approved” whereas this association was not observed for self-citations. This indicates that missing citations that were not self-citations were considered more serious than missing self-citations.

Examining the context of the self-citation, we found vague or non-existent justifications (Supplement S.8), showing that some reviewers ignored the journals’ guidelines to state their reasoning when including self-citations. Reviewers who included self-citations and withheld their approval were more likely to use the words “need” and “please”, indicating use of coercive language. We also found a large increase in reviewers’ approval after they suggested and then received a citation. Some reviewers are not adhering to good scientific practice and instead are treating peer review as a chance to boost their own career.

For both research questions, the effects were stronger for the second and later versions of the article than for the first version. Reviewers may understand that authors may be more willing to compromise on later versions when they are closer to obtaining an indexed publication. Most researchers understand that the peer review system is imperfect and that they must sometimes make compromises to be successful [20, 49].

Potential improvements to peer review

Journals could give stronger guidance to reviewers and authors on coercive citations [4]. However, given the limited time for peer review and the many differences in guidelines between journals [50], most authors may not read peer review instructions. Hence, guidance alone may have limited impact.

One suggestion is that reviewers declare to editors when they have recommended citations to their own work [51]. A useful innovation would be for all reviews that contain self-citations to be automatically flagged to the editors who could check if the self-citations are justified. We are aware of one journal where this is already happening (personal communication, Benno Torgler). F1000 have recently introduced checks to prevent reviewers from publishing a review with three or more self-citations. If the reviewers continue to request more than three, then the review is examined, and if the self-citations are deemed inappropriate and the reviewer declines to remove them, then the review is declined.

Open peer review has been suggested as a way to reduce coercive citations [7, 51]. However, our results from four journals that use open review show that it is not a perfect antidote, although the problem could be worse in journals using blinded peer review. The transparency of open peer review should prevent reviewers from leaving self-serving comments; however, we found some dubious justifications for self-citations and blatant use of AI (Supplement S.6). These reviewers may have rationalised that although their words are public, they are rarely scrutinised (Supplement S.7); hence, it was worth the risk. The assumed additional quality assurance from open peer review [52] may often be absent.

A more radical change to peer review is that the reviewers initially see a version of the article with all references blinded and no reference list; for example, “A strong association between increased cleaning and reduced hospital infection is well established [x]”. Reviewers are asked to give an initial recommendation and comments, and then are shown the version with the full references and asked if they need to update their recommendation or provide additional comments. However, this involves more administrative work and demands more from peer reviewers. This approach could be used for particularly consequential or controversial articles. Some journals already require authors to partially blind their articles to maintain anonymous peer review; for example, the instructions from Taylor & Francis include blinding the authors’ names in the reference list [53].

An argument could be made for using AI to provide peer review that is unmoved by citation flattery. However, peer review is an inherently human task by peers, and instead we need to improve peer review rather than abdicating this often difficult and time-consuming task to machines [54].

Related research

Previous cross-sectional studies of self-citations in reviews found at least one self-citation in 3% at a journal that used blinded peer review [17], 12% at a journal that used blinded peer review [55], and 12% at a journal that used open peer review [56]. A related study found that 15% of reviews included a self-citation and that the self-citations were highest when the reviewer recommended “major revisions” [57]. These figures are comparable with the 6% found here and indicate that most reviews do not include self-citations.

Previous surveys estimated that 14% of authors had experienced a coercive citation request from an editor [58], and 7% and 23% had experienced coercive citation pressure from a reviewer [14, 59]. The frequency with which researchers interact with peer review means that many will encounter coercive citations at some point in their career.

A study of conference submissions estimated that reviewers who were cited gave submissions much higher scores [18]. A study of journal peer review estimated that cited reviewers scored the article higher, but with potential confounding by the quality of the article [17].

A survey of authors concluded that accepting an editor’s request for citations improved the chances of being accepted [12]. Requests in later versions were more strongly associated with acquiescence, and we found a related pattern in our analysis, with reviewers who requested self-citations being much less likely to recommend approval for later versions (Figure 3). A study examining open peer review found that self-citation requests were more likely to be included than other suggested citations, indicating that many authors wanted to please the reviewer or felt pressure to do so [56].

A survey of journal editors found that only 5% objected to reviewers self-citing, and that this should be expected as reviewers are likely to have done related work [8].

A cross-sectional study found that self-citations were more likely to have no rationale compared to other citations, suggesting that they are more likely to be unwarranted [55].

Strengths and limitations

To the best of our knowledge, this is the first analysis to use a matched design when examining reviewer citations, and hence strongly control for any confounding by the characteristics of the authors or articles. We compared reviewers who examined an identical article; hence, the differences we found should be due to the reviewers.

Our models include measurement error, as some citations to the reviewers’ work will be missed by our data collection, and some captured citations will be inaccurate [60]. We performed random data checks that showed a good accuracy (Supplement S.6); however, we also found valid citations that were not captured by our data extraction for conference proceedings and technical reports, which are less likely to have a DOI. This measurement error would most likely underestimate a true association, as it reduced the variance in citation counts and created a regression dilution [61]. Our estimates will be biased if the associations between citations and reviewers’ recommendations are different for publications that do not have a DOI. Reviewers should be equally happy with any citation to their work; however, some reviewers may prefer citations to indexed articles, as these are more likely to count toward their h-indices [62].

We examined whether citing a reviewer altered their recommendation, but did not examine the sentiment of the citation [63]. Some citations would likely have been critical of the reviewer’s work and we would expect these to reduce the chances of a favourable recommendation. An analysis that included the sentiment of the citation would be useful, although previous research found that most citations are neutral or positive [63].

Our results may not be generalisable to journals that use blinded peer review or journals that use the traditional peer review model rather than the publish–review–curate model studied here. A previous study found that asking reviewers to consent to an open review had no important effect on the quality of the review or the reviewers’ recommendation [64]. Another potential difference is that the journals in our sample often asked the authors to suggest peer reviewers; however, this is relatively common in other journals [8].

We found a bias in our sample, as co-reviewers and reviewers from older articles were more likely to be excluded due to not having an OpenAlex record (Supplement S.5). We therefore lost more junior reviewers who were less likely to be cited. The percentage of reviews lost was 5% (2026 out of 39,113), which is hopefully small enough to avoid a large bias.

eLife Assessment

This important study explored a number of issues related to citations in the peer review process. An analysis of more than 37000 peer reviews at four journals found that: i) during the first round of review, reviewers were less likely to recommend acceptance if the article under review cited the reviewer's own articles; ii) during the second and subsequent rounds of review, reviewers were more likely to recommend acceptance if the article cited the reviewer's own articles; iii) during all rounds of review, reviewers who asked authors to cite the reviewer's own articles (a practice known as 'coercive citation') were less likely to recommend acceptance. However, when an author agreed to cite work by the reviewer, the reviewer was more likely to recommend acceptance of the revised article. The evidence is convincing, but article would benefit from a clearer presentation of the results and a more nuanced discussion of the motivations of reviewers.

Reviewer #1 (Public Review):

Summary:

The work used open peer reviews and followed them through a succession of reviews and author revisions. It assessed whether a reviewer had requested the author include additional citations and references to the reviewers' work. It then assessed whether the author had followed these suggestions and what the probability of acceptance was based on the authors decision.

Strengths and weaknesses:

The work's strengths are the in-depth and thorough statistical analysis it contains and the very large dataset it uses. The methods are robust and reported in detail. However, this is also a weakness of the work. Such thorough analysis makes it very hard to read! It's a very interesting paper with some excellent and thought provoking references but it needs to be careful not to overstate the results and improve the readability so it can be disseminated widely. It should also discuss more alternative explanations for the findings and, where possible, dismiss them.

Reviewer #2 (Public Review):

Summary:

This article examines reviewer coercion in the form of requesting citations to the reviewer's own work as a possible trade for acceptance and shows that, under certain conditions, this happens.

Strengths:

The methods are well done and the results support the conclusions that some reviewers "request" self-citations and may be making acceptance decisions based on whether an author fulfills that request.

Weaknesses:

The author needs to be more clear on the fact that, in some instances, requests for self-citations by reviewers is important and valuable.

Reviewer #3 (Public Review):

Summary:

In this article, Barnett examines a pressing question regarding citing behavior of authors during the peer review process. In particular, the author studies the interaction between reviewers and authors, focusing on the odds of acceptance, and how this may be affected by whether or not the authors cited the reviewers' prior work, whether the reviewer requested such citations be added, and whether the authors complied/how that affected the reviewer decision-making.

Strengths:

The author uses a clever analytical design, examining four journals that use the same open peer review system, in which the identities of the authors and reviewers are both available and linkable to structured data. Categorical information about the approval is also available as structured data. This design allows a large scale investigation of this question.

Weaknesses:

My concerns pertain to the interpretability of the data as presented and the overly terse writing style.

Regarding interpretability, it is often unclear what subset of the data are being used both in the prose and figures. For example, the descriptive statistics show many more Version 1 articles than Version 2+. How are the data subset among the different possible methods?

Likewise, the methods indicate that a matching procedure was used comparing two reviewers for the same manuscript in order to control for potential confounds. However, the number of reviews is less than double the number of Version 1 articles, making it unclear which data were used in the final analysis. The methods also state that data were stratified by version. This raises a question about which articles/reviews were included in each of the analyses. I suggest spending more space describing how the data are subset and stratified. This should include any conditional subsetting as in the analysis on the 441 reviews where the reviewer was not cited in Version 1 but requested a citation for Version 2. Each of the figures and tables, as well as statistics provided in the text should provide this information, which would make this paper much more accessible to the reader. [Note from editor: Please see "Editorial feedback" for more on this]

Finally, I would caution against imputing motivations to the reviewers, despite the important findings provided here. This is because the data as presented suggest a more nuanced interpretation is warranted. First, the author observes similar patterns of accept/reject decisions whether the suggested citation is a citation to the reviewer or not (Figs 3 and 4). Second, much of the observed reviewer behavior disappears or has much lower effect sizes depending on whether "Accept with Reservations" is considered an Accept or a Reject. This is acknowledged in the results text, but largely left out of the discussion. The conditional analysis on the 441 reviews mentioned above does support a more cautious version of the conclusion drawn here, especially when considered alongside the specific comments left by reviewers that were mentioned in the results and information in Table S.3. However, I recommend toning the language down to match the strength of the data.

Reviewer #4 (Public Review):

Summary:

This work investigates whether a citation to a referee made by a paper is associated with a more positive evaluation by that referee for that paper. It provides evidence supporting this hypothesis. The work also investigates the role of self citations by referees where the referee would ask authors to cite the referee's paper.

Strengths:

This is an important problem: referees for scientific papers must provide their impartial opinions rooted in core scientific principles. Any undue influence due to the role of citations breaks this requirement. This work studies the possible presence and extent of this.

Barring a few issues discussed below, the methods are solid and well done. The work uses a matched pair design which controls for article-level confounding and further investigates robustness to other potential confounds.

It is surprising that even in these investigated journals where referee names are public, there is prevalence of such citation-related behaviors.

Weaknesses:

Some overall claims are questionable:

"Reviewers who were cited were more likely to approve the article, but only after version 1" It also appears that referees who were cited were less likely to approve the article in version 1. This null or slightly negative effect undermines the broad claim of citations swaying referees. The paper highlights only the positive results while not including the absence (and even reversal) of the effect in version 1 in its narrative.

"To the best of our knowledge, this is the first analysis to use a matched design when examining reviewer citations" Does not appear to be a valid claim based on the literature reference [18]

It will be useful to have a control group in the analysis associated to Figure 5 where the control group comprises matched reviews that did not ask for a self citation. This will help demarcate words associated with approval under self citation (as compared to when there is no self citation). The current narrative appears to suggest an association of the use of these words with self citations but without any control.

More discussion on the recommendations will help: For the suggestion that "the reviewers initially see a version of the article with all references blinded and no reference list" the paper says "this involves more administrative work and demands more from peer reviewers". I am afraid this can also degrade the quality of peer review, given that the research cannot be contextualized properly by referees. Referees may not revert back to all their thoughts and evaluations when references are released afterwards.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review)::

Summary:

The work used open peer reviews and followed them through a succession of reviews and author revisions. It assessed whether a reviewer had requested the author include additional citations and references to the reviewers' work. It then assessed whether the author had followed these suggestions and what the probability of acceptance was based on the authors decision.

Strengths and weaknesses:

The work's strengths are the in-depth and thorough statistical analysis it contains and the very large dataset it uses. The methods are robust and reported in detail. However, this is also a weakness of the work. Such thorough analysis makes it very hard to read! It's a very interesting paper with some excellent and thought provoking references but it needs to be careful not to overstate the results and improve the readability so it can be disseminated widely. It should also discuss more alternative explanations for the findings and, where possible, dismiss them.

I have toned down the language including a more neutral title. To help focus on the main results, I have moved four paragraphs from the methods to the supplement. These are the sample size, the two sensitivity analyses on including co-reviewers and confounding by reviewers’ characteristics, and the analysis examining potential bias for the reviewers with no OpenAlex record.

Reviewer #2 (Public review):

Summary:

This article examines reviewer coercion in the form of requesting citations to the reviewer's own work as a possible trade for acceptance and shows that, under certain conditions, this happens.

Strengths:

The methods are well done and the results support the conclusions that some reviewers "request" self-citations and may be making acceptance decisions based on whether an author fulfills that request.

Weaknesses:

The author needs to be more clear on the fact that, in some instances, requests for selfcitations by reviewers is important and valuable.

This is a key point. I have included a new text analysis to examine this issue and have addressed this in the updated discussion.

Reviewer #3 (Public review):

Summary:

In this article, Barnett examines a pressing question regarding citing behavior of authors during the peer review process. In particular, the author studies the interaction between reviewers and authors, focusing on the odds of acceptance, and how this may be affected by whether or not the authors cited the reviewers' prior work, whether the reviewer requested such citations be added, and whether the authors complied/how that affected the reviewer decision-making.

Strengths:

The author uses a clever analytical design, examining four journals that use the same open peer review system, in which the identities of the authors and reviewers are both available and linkable to structured data. Categorical information about the approval is also available as structured data. This design allows a large scale investigation of this question.

Weaknesses:

My concerns pertain to the interpretability of the data as presented and the overly terse writing style.

Regarding interpretability, it is often unclear what subset of the data are being used both in the prose and figures. For example, the descriptive statistics show many more Version 1 articles than Version 2+. How are the data subset among the different possible methods?

I have now included the number of articles and reviews in the legends of each plot. There are more version 1 articles because some are “approved” at this stage and hence a second version is never submitted (I’ve now specifically mentioned this in the discussion).

Likewise, the methods indicate that a matching procedure was used comparing two reviewers for the same manuscript in order to control for potential confounds. However, the number of reviews is less than double the number of Version 1 articles, making it unclear which data were used in the final analysis. The methods also state that data were stratified by version. This raises a question about which articles/reviews were included in each of the analyses. I suggest spending more space describing how the data are subset and stratified. This should include any conditional subsetting as in the analysis on the 441 reviews where the reviewer was not cited in Version 1 but requested a citation for Version 2. Each of the figures and tables, as well as statistics provided in the text should provide this information, which would make this paper much more accessible to the reader.

[Note from editor: Please see "Editorial feedback" for more on this]

The numbers are now given in every figure legend, and show the larger sample size for the first versions.

The analysis of the 441 reviews was an unplanned analysis that is separate to the planned models. The sample size is much smaller than the main models due to the multiple conditions applied to the reviewers: i) reviewed both versions, ii) not cited in first version, iii) requested a self-citation in their first review.

Finally, I would caution against imputing motivations to the reviewers, despite the important findings provided here. This is because the data as presented suggest a more nuanced interpretation is warranted. First, the author observes similar patterns of accept/reject decisions whether the suggested citation is a citation to the reviewer or not (Figs 3 and 4). Second, much of the observed reviewer behavior disappears or has much lower effect sizes depending on whether "Accept with Reservations" is considered an Accept or a Reject. This is acknowledged in the results text, but largely left out of the discussion. The conditional analysis on the 441 reviews mentioned above does support a more cautious version of the conclusion drawn here, especially when considered alongside the specific comments left by reviewers that were mentioned in the results and information in Table S.3. However, I recommend toning the language down to match the strength of the data.

I have used more cautious language throughout, including a new title. The new text analysis presented in the updated version also supports a more cautious approach.

Reviewer #4 (Public review):

Summary:

This work investigates whether a citation to a referee made by a paper is associated with a more positive evaluation by that referee for that paper. It provides evidence supporting this hypothesis. The work also investigates the role of self citations by referees where the referee would ask authors to cite the referee's paper.

Strengths:

This is an important problem: referees for scientific papers must provide their impartial opinions rooted in core scientific principles. Any undue influence due to the role of citations breaks this requirement. This work studies the possible presence and extent of this.

Barring a few issues discussed below, the methods are solid and well done. The work uses a matched pair design which controls for article-level confounding and further investigates robustness to other potential confounds.

It is surprising that even in these investigated journals where referee names are public, there is prevalence of such citation-related behaviors.

Weaknesses:

Some overall claims are questionable:

"Reviewers who were cited were more likely to approve the article, but only after version 1" It also appears that referees who were cited were less likely to approve the article in version 1. This null or slightly negative effect undermines the broad claim of citations swaying referees. The paper highlights only the positive results while not including the absence (and even reversal) of the effect in version 1 in its narrative.

The reversed effect for version 1 is interesting, but the adjusted 99.4% confidence interval includes 1 and hence it’s hard to be confident that this is genuinely in the reverse direction. However, it is certainly far from the strongly positive association for versions 2+.

"To the best of our knowledge, this is the first analysis to use a matched design when examining reviewer citations" Does not appear to be a valid claim based on the literature reference [18]

This previous paper used a matched design but then did not used a matched analysis. Hence, I’ve changed the text in my paper to “first analysis to use a matched design and analysis”. This may seem a minor claim of novelty, but not using a matched analysis for matched data could discard much of the benefits of the matching.

It will be useful to have a control group in the analysis associated to Figure 5 where the control group comprises matched reviews that did not ask for a self citation. This will help demarcate words associated with approval under self citation (as compared to when there is no self citation). The current narrative appears to suggest an association of the use of these words with self citations but without any control.

Thanks for this useful suggestion. I have added a control group of reviewers who requested citations to articles other than their own. The words requested were very similar to the previous analysis, hence I’ve needed to reinterpret the results from the text analysis as “please” and “need” are not exclusively used by those requesting selfcitations. I also fixed a minor error in the text analysis concerning the exclusion of abstracts of shorter than 100 characters.

More discussion on the recommendations will help:

For the suggestion that "the reviewers initially see a version of the article with all references blinded and no reference list" the paper says "this involves more administrative work and demands more from peer reviewers". I am afraid this can also degrade the quality of peer review, given that the research cannot be contextualized properly by referees. Referees may not revert back to all their thoughts and evaluations when references are released afterwards.

This is an interesting point, but I don’t think it’s certain that this would happen. For example, revisiting the review may provide a fresh perspective and new ideas; this sometimes happens for me when I review the second version of an article. Ideally an experiment is needed to test this approach, as it is difficult to predict how authors and reviewers will react.

Recommendations for the Authors:

Editorial feedback:

I wonder if the article would benefit from a shorter title, such as the one suggested below. However, please feel free to not change the title if you prefer.

[i] Are peer reviewers influenced by their work being cited (or not)?

I like the slightly simpler: “Are peer reviewers influenced by their work being cited?”

[ii] To better reflect the findings in the article, please revise the abstract along the following lines:

Peer reviewers for journals sometimes write that one or more of their own articles should have been cited in the article under review. In some cases such comments are justified, but in other cases they are not. Here, using a sample of more than 37000 peer reviews for four journals that use open peer review and make all article versions available, we use a matched study design to explore this and other phenomena related to citations in the peer review process. We find that reviewers who were cited in the article under review were less likely to approve the original version of an article compared with reviewers who were not cited (odds ratio = 0.84; adjusted 99.4% CI: 0.69-1.03), but were more likely to approve a revised article in which they were cited (odds ratio = 1.61; adjusted 99.4% CI: 1.16-2.23). Moreover, for all versions of an article, reviewers who asked for their own articles to be cited were much less likely to approve the article compared with reviewers who did not do this (odds ratio = 0.15; adjusted 99.4% CI: 0.08-0.30). However, reviewers who had asked for their own articles to be cited were much more likely to approve a revised article that cited their own articles compared to a revised article that did not (odds ratio = 3.5; 95% CI: 2.0-6.1).

I have re-written the abstract along the lines suggested. I have not included the finding that cited reviewers were less likely to approve the article due to the adjusted 99.4% interval including 1.

[iii] The use of the phrase "self-citation" to describe an author citing an article by one of the reviewers is potentially confusing, and I suggest you avoid this phrase if possible.

I have removed “self-citation” everywhere and instead used “citations to their own articles”.

[iv] I think the captions for figures 2, 3 and 4 from benefit from rewording to more clearly describe what is being shown in the figure. Please consider revising the caption for figure 2 as follows, and revising the captions for figures 3 and 4 along similar lines. Please also consider replotting some of the panels so that the values on the horizontal axes of the top panel align with the values on the bottom panel.

I have aligned the odds and probability axes as suggested which better highlights the important differences. I have updated the figure captions as outlined.

Figure 2: Odds ratios and probabilities for reviewers giving a more or less favourable recommendation depending on whether they were cited in the article.

Top left: Odds ratios for reviewers giving a more favourable (Approved) or less favourable (Reservations or Not approved) recommendation depending on whether they were cited in the article. Reviewers who were cited in version 1 of the article (green) were less likely to make a favourable recommendation (odds ratio = 0.84; adjusted 99.4% CI: 0.691.03), but they were more likely to make a favourable recommendation (odds ratio = 1.61; adjusted 99.4% CI: 1.16-2.23) if they were cited in a subsequent version (blue). Top right: Same data as top left displayed in terms of probabilities. From the top, the lines show the probability of a reviewer approving: a version 1 article in which they are not cited (please give mean value and CI); a version 1 article in which they are cited (mean value and CI); a version 2 (or higher) article in which they are not cited (mean value and CI); and a version 2 (or higher) article in which they are cited (mean value and CI).

Bottom left: Same data as top left except that more favourable is now defined as Approved or Reservations, and less favourable is defined as Not approved. Again, reviewers who were cited in version 1 were less likely to make a favourable recommendation (odds ratio = 0.84; adjusted 99.4% CI: 0.57-1.23),and reviewers who were cited in subsequent versions were more likely to make a favourable recommendation (odds ratio = 1.12; adjusted 99.4% CI: 0.59-2.13).

Bottom right: Same data as bottom left displayed in terms of probabilities. From the top, the lines show the probability of a reviewer approving: a version 1 article in which they are not cited (please give mean value and CI); a version 1 article in which they are cited (mean value and CI); a version 2 (or higher) article in which they are not cited (mean value and CI); and a version 2 (or higher) article in which they are cited (mean value and CI).

This figure is based on an analysis of [Please state how many articles, reviewers, reviews etc are included in this analysis].

In all the panels a dot represents a mean, and a horizontal line represents an adjusted 99.4% confidence interval.

Reviewer #1 (Recommendations for the Authors):

A big recommendation to the author would be to consider putting a lot of the statistical analysis in an appendix and describing the methods and results in more accessible terms in the main text. This would help more readers see the baby through the bath water

I have moved four paragraphs from the methods to the supplement. These are the sample size, the two sensitivity analyses on including co-reviewers and confounding by reviewers’ characteristics, and the analysis examining potential bias for the reviewers with no OpenAlex record.

One possibility, that may have been accounted for, but it is hard to say given the density of the analysis, is the possibility that an author who follows the recommendations to cite the reviewer has also followed all the other reviewer requests. This could account for the much higher likelihood of acceptance. Conversely an author who has rejected the request to cite the reviewer may be more likely to have rejected many of the other suggestions leading to a rejection. I couldn't discern whether the analysis had accounted for this possibility. If it has it need to be said more prominently, if it hasn't this possibility at least needs to be discussed. It would be good to see other alternative explanations for the results discussed (and if possible dismissed) in the discussion section too.

This is an interesting idea. It’s also possible that authors more often accept and include any citation requests as it gives them more license to push back on other more involved changes that they would prefer not to make, e.g., running a new analysis. To examine this would require an analysis of the authors’ responses to the reviewers, and I have now added this as a limitation.

I hope this paper will have an impact on scientific publishing but I fear that it won't. This is no reflection on the paper but a more a reflection on the science publishing system.

I do not have any additional references (written by myself or others!) I would like the author to include

Thanks. I appreciate that extra thought is needed when peer reviewing papers on peer review. I do not know the reviewers’ names! I have added one additional reference suggested by the reviewers which had relevant results on previous surveys of coercive citations for the section on “Related research”.

Reviewer #2 (Recommendations for the Authors):

(1) Would it be possible for the author to control for academic discipline? Some disciplines cite at different rates and have different citation sub-cultures; for example, Wilhite and Fong (2012) show that editorial coercive citation differs among the social science and business disciplines. Is it possible that reviewers from different disciplines just take a totally different view of requesting self-citations?

Wilhite, A.W., & Fong, E.A. 2012. Coercive citation in academic publishing. Science, 335: 542-543.

This is an interesting idea, but the number of disciplines would need to be relatively broad to keep a sufficient sample size. The Catch-22 is then whether broad disciplines are different enough to show cultural differences. Overall, this is an idea for future work.

(2) I would like the author to be much more clear about their results in the discussion section. In line 214, they state that "Reviewers who requested a self-citation were much less likely to approve the article for all versions." Maybe in the discussion some language along the lines of "Although reviewers who requested self-citation were actually much less likely to approve an article, my more detailed analyses show that this was not the case when reviewers requested a self-citation without reason or with the inclusion of coercive language such as 'need' or 'please'." Again, word it as you like, but I think it should be made clear that requests for self-citation alone is not a problem. In fact, I would argue that what the author says in lines 250 to 255 in the discussion reflects that reviewers who request self-citations (maybe for good reasons) are more likely to be the real experts in the area and why those who did not request a self-cite did not notice the omission. It is my understanding that editors are trying to get warm bodies to review and thus reviewers are not all equally qualified. Could it be that requesting self-citations for a good reason is a proxy for someone who actually knows the literature better? I'm not saying this is s fact, but it is a possibility. I get this is said in the abstract, but worth fleshing out in the discussion.

I have updated the discussion after a new text analysis and have addressed this important question of whether self-citations are different from citations to other articles. The idea that some self-citers are more aware of the relevant literature is interesting, although this is very hard to test because they could also just be more aware of their own work. The question of whether self-citations are justified is a key question and one that I’ve tried to address in an updated discussion.

Reviewer #3 (Recommendations for the Authors):

Data and code availablility are in good shape. At a high level, I recommend:

Toning down the interpretation of reviewers' motivation, especially since some of this is mitigated by findings presented in the paper.

I have reworded the discussion and included a warning on the observational study design.

Devote more time detailing exactly what data are being presented in each figure/table and results section as described in more detail in the main review (n, selection criteria, conditional subsetting, etc.).

I agree and have provided more details in each figure legend.

Reviewer #4 (Recommendations for the Authors):

A few aspects of the paper are not clear:

I did not follow Figure 4. Are the "self citation" labels supposed to be "citation to other research"?

Thanks for picking up this error which has now been fixed.

I did not understand how to parse the left column of Figure 2

As per the editor’s suggestion, the figure legend has been updated.

Table 3: Please use different markers for the different curves so that it is clearly demarcated even in grayscale print

I presume you meant Figure 3 not Table 3. I’ve varied the symbols in all three odds ratio plots.

Supplementary S3: Typo "Approvep" Fixed, thanks.

OTHER CHANGES: As well as the four reviews, my paper was reviewed by an AI-reviewer which provided some useful suggestions. I have mentioned this review in the acknowledgements. I have reversed the order of figure 5 to show the probability of “Approved” as this is simpler to interpret.

Version 2

Abstract

Peer reviewers sometimes comment that their own journal articles should be cited by the journal article under review. Comments concerning relevant articles can be justified, but comments can also be unrelated coercive citations. Here, we used a matched study design to explore how citations influence the peer review process. We used a sample of more than 37,000 peer reviews from four journals that use open peer review and make all article versions available. We find that reviewers who were cited in the article under review were more likely to recommend approval, but only after the first version (odds ratio = 1.61; adjusted 99.4% CI: 1.16 to 2.23). For all versions of the articles, the reviewers who commented that their own articles should be cited were less likely to recommend approval compared to the reviewers who did not, with the strongest association after the first version (odds ratio = 0.15; adjusted 99.4% CI: 0.08 to 0.30). Reviewers who included a citation to their own articles were much more likely to approve a revised article that cited their articles compared to a revised article that did not (odds ratio = 3.5; 95% CI: 2.0 to 6.1). Some reviewers’ recommendations depend on whether they are cited or want to be cited. Reviewer citation requests can turn peer review into a transaction rather than an objective critique of the article.

Introduction

In 2024, a published peer reviewed article included this remarkable sentence: “As strongly requested by the reviewers, here we cite some references [35–47] although they are completely irrelevant to the present work” [1]. This was a rare public example of coerced citations, where a reviewer exploits the peer review process to increase their citation counts and hence further their own career [2–4]. Reviewers should be relevant experts, so some suggestions to cite their articles will be appropriate. However, excessive citation requests or requests to cite unrelated articles are unethical [5–9]. Coerced citations can also come from editors trying to boost their journal’s ranking [10–12].

Coerced citations are reported as a common problem in peer review. In author surveys, two-thirds reported pressure from peer reviewers to cite unrelated articles [13] and 23% had experienced a reviewer that “required them to include unnecessary references to their publication(s)” [14]. Publishers have investigated whether “hundreds of researchers” have manipulated the peer review process to increase their own citations [15]. Some reviewers may be exploiting their power over authors who “have a strong incentive to [...] accept all ‘suggestions’ by the referees even if one knows that they are misleading or even incorrect” [16].

As reviewers are often in the same field as the article’s authors, they may already be cited in the article without the need for coerced citations. Reviewers who are cited may give a more favourable peer review and be more willing to overlook flaws [17, 18]. Some authors may try to exploit this using “referee baiting” [3] or “flattery citations” [19] by favourably citing a reviewer’s work.

The interactions during peer review between authors and reviewers can determine whether an article is accepted [20] and what results are included in the published version [21]. Given the importance of peer review for science, studies that examine how peer review works in practice are needed [22–26]. Here, we examine interactions between peer reviewers and authors using four journals that publish all article versions and all peer reviews. We had two research questions:

Do peer reviewers give a more or less favourable recommendation when they are cited in the article?

Do peer reviewers give a more or less favourable recommendation when their review includes a citation to their own articles?

Methods

Journal selection

We studied journals from the publisher F1000 as their journals use open peer review with signed reviewers. F1000 journals use a publish–review–curate model [27], meaning all versions of the article are publicly available, including versions updated after peer review. This allowed us to examine the interactions between authors and reviewers throughout the peer review process. We selected four F1000 journals that each had over 100 articles. Some characteristics of the four journals are given in Table 1. Three journals were created to support funders.

The peer review process used by F1000 journals differs from most standard journals. The journals do not use academic editors, but do have in-house editors who manage articles but do not make editorial decisions. This means that most interactions during peer review are between authors and reviewers directly. In-house editors perform checks prior to the first version of the article being published and at F1000Research this results in 40 to 50% submissions being rejected (personal communication, F1000 staff). Up to mid-2024, authors were asked to identify potential reviewers who were qualified experts with no competing interests [28]. Since mid-2024, reviewer identification is made in-house, although authors can suggest reviewers.

Reviewers are asked to recommend one of three categories: Approved, Approved with reservations, and Not approved. For brevity, we refer to ‘Approved with Reservations’ as ‘Reservations’. An article is indexed once it receives two ‘Approved’ or two ‘Reservations’ and one ‘Approved’. The guidelines for recommending Approved are: “the aims and research methods are adequate; results are presented accurately, and the conclusions are justified and supported by the presented data” [29]. Peer reviewers are asked to assess the validity of an article’s content, rather than novelty or interest levels, an approach designed to combat publication bias [30].

All four journals have a peer reviewer code of conduct and state that reviewers should familiarise themselves with the ethical guidelines for peer reviewers by the Committee On Publication Ethics [31]. The journals’ guidelines for reviewers include the following: “reviewers should explicitly state their reasoning when asking authors to cite their own work”.

Data extraction

We extracted data on authors and articles from the OpenAlex database (https://openalex.org/) and directly from the four journals. OpenAlex combines scholarly data from multiple sources, including ORCID – a unique identifier for researchers, Microsoft Academic, Crossref and PubMed. A recent study compared OpenAlex with the two commonly used proprietary bibliometric databases of Web of Science and Scopus for the years 2015 to 2022 [32]. The results were mixed, but OpenAlex had better ORCID coverage and covered more Digital Object Identifiers (DOIs) – the unique identifier for publications. We accessed OpenAlex using the openalexR package [33, 34]. We used each journal’s application programming interface (API) to extract data on the articles and peer reviews. The data were extracted in four stages:

Searches were made using the APIs of the four journals to find all articles published between 1 Jan 2012 and 28 May 2025, with the start date to capture all potential articles.

For each article, the following data were downloaded in XML format:

The article’s publication date and version number

The reviewers’ names and ORCIDs (if available)

The text of all reviews and the reviewers’ recommendations

The DOIs and PMIDs (PubMed IDs) from the article’s reference list

The DOIs and PMIDs of any articles cited by the reviewers. The online peer review system at F1000 journals includes the DOI of any article cited in the review, which facilitates the identification of citations to the reviewers’ articles.

Articles were excluded if:

They were not peer reviewed or had yet to receive any reviews

The reference list was empty

The reviewers’ publication histories were collected from OpenAlex using their name, institution and ORCID (if available). Reviews were excluded if there was no record for the reviewer in OpenAlex, or if the reviewer had no published articles as there was no potential for them to be cited or request a citation to their own articles.

Study design

We used two predictor variables about the reviewer:

The number of times they were cited in the article (0, 1, 2, …).

The number of times they included citations to their own articles in their review (0, 1, 2, …).

We fitted both predictors as linear, but reviewers may behave differently with any citation rather than a linear change, and hence we also fitted both predictors as a binary “none versus any” (0 versus 1, 2, …). We compared the linear and binary alternatives using the Akaike Information Criterion (AIC) to find the parameterisation that best fitted the data [35].

We matched on article and version to control for confounding by any characteristics of the article [36]; for example, the article’s topic or writing style. Hence, we compared two or more independent reviewers who considered the same article.

All analyses were stratified by article version, using the first version only or the second and subsequent versions. This is because the reviewers are unknown to the authors for the first version, but from the second version onwards, the authors will know the reviewers as the journals use signed peer reviews. This knowledge could alter the behaviour of authors and reviewers.

The study design is summarised in Figure 1

Graphical summary of the study design for research question 1 showing a dummy article and two reviews.

In the first version of the article, the reviewer Smith (blue) is cited whilst Jones (purple) is not. For the second version of the article, the authors are now aware that Jones is a reviewer and Jones has been cited. The reviewers’ recommendations are the outcome and are colour-coded as Not approved (red), Reservations (orange) and Approved (green). We tested whether citations to the reviewer in the article influenced their recommendation. The matched design means that only reviewers of the same article are compared (here, Smith and Jones) and the overall effect is estimated by aggregating over multiple matched comparisons. Research question 2 used the same design but examined citations to the reviewers’ articles in their reviews.

Statistical methods

We used conditional logistic regression to examine the associations between the citations to the reviewer and their ordinal recommendation (Approved → Reservations → Not approved) while matching the article and the version [37]. Conditional logistic regression requires a binary dependent variable; hence, we fitted two related models that examined the odds of:

“Approved” compared with “Reservations” or “Not approved”.

“Approved” or “Reservations” compared with “Not approved”,

These two models tested the same hypothesis, hence we adjusted for multiple testing. We also used repeated testing due to the stratification by article version and the two formulations of the predictors (linear or none versus any). Since we used 8 (2 × 2 × 2) tests, we displayed all the results using 99.4% confidence intervals instead of 95.0% intervals, which is a 5% type I error divided by eight tests.

In an unplanned analysis, we examined the association between the reviewer’s recommendation and whether they included citations to work other than citations to their own articles. This was added to examine differences between reviewers’ citations to their own articles and other articles.

Outliers were not excluded. No data were missing in the analysis data set.

The sample size calculation is in Supplement S.1.

Text analysis

We examined how reviewers’ citations to their own articles or other articles were justified and whether their wording differed according to their recommendation. For an initial view of citations to their own articles, we randomly selected 20 reviews and extracted the most relevant sentence concerning the citation.

To analyse the review text, we first extracted the 100 most commonly used words in all reviews. To standardise the text, all words were transformed into tokens, with stop-words removed and then stemmed. We then tested which of the 100 words were associated with recommending Approved versus Reservations or Not approved amongst those reviewers who included a citation to their own articles and those who included a citation to other articles. We chose the set of words using an elastic net with 10-fold cross-validation and selected a parsimonious model by using the lambda within one standard error of the minimum cross-validated error [38, 39]. To get uncertainty intervals for the estimates, we fitted a Bayesian model with the set of words selected by the elastic net and using a sceptical Normal prior centred on zero to create shrinkage.

Reproducibility

Research question 1 was pre-registered using As Predicted on 20 May 2024 [40]. Research question 2 was formulated during data collection but before any data analysis and used the same study design and statistical methods as question 1.

All data extraction and analyses were conducted using R version 4.4.1 [41]. The data and R code are available on GitHub [42].

Results

A flow chart of the included reviews is shown in Supplement S.2. The final sample size was over 37,000 reviews. There were more than 3,500 articles that were not included because they had not yet been peer reviewed, especially recent articles. More than 2,000 reviewers did not have a record in OpenAlex and so could not be included. These missing reviewers were more likely to be from older articles and more likely to be co-reviewers.

Descriptive statistics on the included reviews are in Table 2. The reviewers were cited at least once in 13% of the articles and 6% of the reviews included a self-citation. Most reviews recommended “Approved” (54%), with only 8% recommending “Not approved” which is low compared with many journals; however, 40 to 50% of submissions are rejected before articles are sent for peer review (personal communication, F1000 staff). The reviewers were relatively experienced, with a median number of papers of 55.

The binary predictor for citations of “any versus none” had a generally better fit to the data compared to the linear predictor (Supplement S.3). This indicates that for most reviewers receiving any citation is important, and there is no linear increase for two or more citations. The following results are for the binary predictor “any versus none”, with the results using a linear predictor in Supplement S.4.

Reviewers who were cited were more likely to approve the article, but only after version 1 (Fig 2 and Table 3). If a reviewer was cited in any versions after version 1, the odds ratio for recommending Approved versus Reservations or Not approved was 1.61 (adjusted 99.4% CI 1.16 to 2.23).

Odds ratios and probabilities for reviewers giving a more or less favourable recommendation depending on whether they were cited in the article.

Top left: Odds ratios for reviewers giving a more favourable (Approved) or less favourable (Reservations or Not approved) recommendation depending on whether they were cited in the article. Reviewers cited in later versions (blue) were more likely to make a favourable recommendation (odds ratio = 1.61; adjusted 99.4% CI: 1.16 to 2.23), whereas being cited in the first version (green) did not improve their recommendation (odds ratio = 0.84; adjusted 99.4% CI: 0.69 to 1.03). Top right: Same results as top left displayed as conditional probabilities. From the top, the lines show the within-strata probability of a reviewer approving: a version 1 article in which they are not cited (0.51; adjusted 99.4% CI: 0.49 to 0.52); a version 1 article in which they are cited (0.46; adjusted 99.4% CI: 0.41 to 0.51); a version 2 (or higher) article in which they are not cited (0.48; adjusted 99.4% CI: 0.45 to 0.51); and a version 2 (or higher) article in which they are cited (0.60; adjusted 99.4% CI: 0.53 to 0.65). Bottom left: Same estimates as top left except that a more favourable recommendation is now Approved or Reservations and a less favourable is Not approved. There was no clear association for cited reviewers in version 1 (odds ratio = 0.84; adjusted 99.4% CI: 0.57 to 1.23) or later versions (odds ratio = 1.12; adjusted 99.4% CI: 0.59 to 2.13). Bottom right: Same results as bottom left displayed as conditional probabilities. From the top, the lines show the within-strata probability of a reviewer approving: a version 1 article in which they are not cited (0.51; adjusted 99.4% CI: 0.48 to 0.53); a version 1 article in which they are cited (0.47; adjusted 99.4% CI: 0.35 to 0.55); a version 2 (or higher) article in which they are not cited (0.50; adjusted 99.4% CI: 0.43 to 0.55); and a version 2 (or higher) article in which they are cited (0.53; adjusted 99.4% CI: 0.34 to 0.63). This figure is based on an analysis of 12,051 articles and 24,677 reviews for version 1 and 6090 articles and 10,196 reviews for version 2+. In all panels a dot or square represents a mean, and a horizontal line represents an adjusted 99.4% confidence interval.

Reviewers who included a citation to their own articles were much less likely to approve the article for all versions (Fig 3 and Table 3). The odds ratio for recommending Approved versus Reservations or Not approved was 0.57 (99.4% CI 0.44 to 0.73) for version 1 and strengthened to 0.15 (99.4% CI 0.08 to 0.30) for versions 2+. The less favourable recommendation was only for the approval of the article and the odds ratios for Approved or Reservations versus Not approved were much closer to 1.

Odds ratios and probabilities for reviewers giving a more or less favourable recommendation if they included a citation to their own articles in their review.

Top left: Odds ratios for reviewers giving a more favourable (Approved) or less favourable (Reservations or Not approved) recommendation depending on whether their review included a citation to their own articles. Reviewers including a citation to their own articles were less likely to make a favourable recommendation for version 1 (green; odds ratio = 0.57; adjusted 99.4% CI: 0.44 to 0.73) and later versions (blue; odds ratio = 0.15; adjusted 99.4% CI: 0.07 to 0.30). Top right: Same results as top left displayed as conditional probabilities. From the top, the lines show the within-strata probability of a reviewer approving: a version 1 article in which their review did not include a citation (0.51; adjusted 99.4% CI: 0.50 to 0.53); a version 1 article in which in which their review included a citation (0.37; adjusted 99.4% CI: 0.29 to 0.44); a version 2 (or higher) article in which their review did not include a citation (0.51; adjusted 99.4% CI: 0.49 to 0.53); and a version 2 (or higher) article in which in which their review included a citation (0.14; adjusted 99.4% CI: 0.01 to 0.30). Bottom left: Same estimates as top left except that a more favourable recommendation is now Approved or Reservations and a less favourable is Not approved. There was no clear association for reviewers who included a citation to their own articles in version 1 (odds ratio = 1.11; adjusted 99.4% CI: 0.77 to 1.60) or later versions (odds ratio = 0.80; adjusted 99.4% CI: 0.37 to 1.74). Bottom right: Same results as bottom left displayed as conditional probabilities. From the top, the lines show the within-strata probability of a reviewer approving: a version 1 article in which their review did not include a citation (0.50; adjusted 99.4% CI: 0.48 to 0.52); a version 1 article in which their review included a citation (0.53; adjusted 99.4% CI: 0.43 to 0.60); a version 2 (or higher) article in which their review did not include a citation (0.51; adjusted 99.4% CI: 0.46 to 0.54); and a version 2 (or higher) article in which their included a citation (0.45; adjusted 99.4% CI: 0.12 to 0.61). This figure is based on an analysis of 12,078 articles and 24,732 reviews for version 1 and 6101 articles and 10,213 reviews for version 2+. In all panels a dot or square represents a mean, and a horizontal line represents an adjusted 99.4% confidence interval.

In an unplanned analysis, we examined the behaviour of reviewers in the first two versions of the article. We examined the 441 reviews where the reviewer was not cited in version 1 of the article and included a citation to their own articles in their first review. The reviewers who were then cited in version 2 recommended approval for 92% compared to 76% for reviewers who were not cited (odds ratio = 3.5, 95% CI: 2.0 to 6.1). This analysis did not use matching.

In an unplanned analysis, we examined whether the reviewers’ recommendations depended on whether their review included citations to articles other than their own. Reviewers who included citations in their review were much more likely not to approve the article (Figure 4), which was similar to the association with citations to reviewers’ own articles (Figure 3). However, reviewers who included citations to articles other than their own were also much more likely to recommend “Not approved”, as shown by the lower odds of “Approved” or “Reservations” versus “Not approved”. This association was not seen using citations to reviewers’ own articles (Figure 3)

Odds ratios and probabilities for reviewers giving a more or less favourable recommendation depending on if they included citations to articles other than their own in their review.

Top left: Odds ratios for reviewers giving a more favourable (Approved) or less favourable (Reservations or Not approved) recommendation depending on whether their review included a citation to articles other than their own. Reviewers including citations to other articles were less likely to make a favourable recommendation for version 1 (green; odds ratio = 0.53; adjusted 99.4% CI: 0.44 to 0.64) and later versions (blue; odds ratio = 0.18; adjusted 99.4% CI: 0.10 to 0.30). Top right: Same results as top left displayed as conditional probabilities. From the top, the lines show the within-strata probability of a reviewer approving: a version 1 article in which their review did not cite other articles (0.53; adjusted 99.4% CI: 0.51 to 0.54); a version 1 article in which their review cited other articles (0.37; adjusted 99.4% CI: 0.31 to 0.42); a version 2 (or higher) article in which their review did not cite other articles (0.52; adjusted 99.4% CI: 0.50 to 0.54); and a version 2 (or higher) article in which in which their review cited other articles (0.17; adjusted 99.4% CI: 0.02 to 0.30). Bottom left: Same estimates as top left except that a more favourable recommendation is now Approved or Reservations and a less favourable is Not approved. Reviewers including citations to other articles were less likely to make a favourable recommendation for version 1 (odds ratio = 0.62; adjusted 99.4% CI: 0.46 to 0.84) and later versions (odds ratio = 0.34; adjusted 99.4% CI: 0.16 to 0.73). Bottom right: Same results as bottom left displayed as conditional probabilities. From the top, the lines show the within-strata probability of a reviewer approving: a version 1 article in which their review did not cite other articles (0.52; adjusted 99.4% CI: 0.49 to 0.54); a version 1 article in which their review cited other articles (0.41; adjusted 99.4% CI: 0.31 to 0.48); a version 2 (or higher) article in which their review did not cite other articles (0.52; adjusted 99.4% CI: 0.47 to 0.55); and a version 2 (or higher) article in which their review cited other articles (0.27; adjusted 99.4% CI: 0.02 to 0.45). This figure is based on an analysis of 12,078 articles and 24,732 reviews for version 1 and 6101 articles and 10,213 reviews for version 2+. In all panels a dot or square represents a mean, and a horizontal line represents an adjusted 99.4% confidence interval.

Sensitivity analyses

The odds ratios when including co-reviewers with reviewers were similar to the odds ratios when using reviewers only (Supplement S.5).

We found no evidence that the reviewers’ publication numbers or country confounded the associations between citations and recommendations (Supplement S.6).

Text analyses of reviewers’ comments

A random sample of how reviewers included citations to their own articles found some vague justifications (Supplement S.7); for example, “Here are some additional publications you might consider referencing”. Other sentences adhered to the publisher’s guidelines for reviewers, as specific reasoning was provided for citations to their own articles [7]. One reviewer thanked the authors for a previous citation. Three reviews did not have a relevant sentence. One reviewer almost certainly used AI to write their review as it included the phrase “Certainly! Here are some potential review questions for the manuscript” [43]; this review included six self-citations with no justifications.

Reviewers who included a citation to their own articles or other articles were more likely to use the words “need” and “please” when not approving the article (Figure 5). In contrast, the words “genome” and “well” were the most strongly associated with the reviewers’ approval.

Words in the reviewers’ comments that were associated with approving the article or not for reviewers who included a citation to their own articles (n = 2, 025) and reviewers who included citations to other articles (n = 4, 350).

The words were selected using an elastic net that started with the 100 most commonly used review words. The estimates from the elastic net are shown as empty circles and the mean estimates and 95% credible intervals from a Bayesian model are shown as a solid circle and horizontal line. Words are shown if the probability of a non-zero mean was over 0.95 for either reviewers who cited their own articles or reviewers who cited other articles. Four words were selected by the elastic net for the reviewers who cited other articles but not by the elastic net for reviewers who cited their own articles. The axis label shows the stemmed word and most common whole word in brackets.

To examine how often open peer reviews were viewed, we took a random sample of 200 reviews from the four journals and found that, on average, they were viewed just 1.2 times per year (Supplement S.8).

Discussion

Our results provide evidence that some reviewers have a transactional view of peer review, with their final approval dependent on citations to their work. Some reviewers may be exploiting the pressure on authors to “publish or perish”. Under this pressure, many authors may oblige and add the suggested citations, especially since adding another citation may only require a minor edit to their article [44]. Both sides gain from this transaction, as the authors get an indexed publication and the reviewer gets a citation.

A key question is whether citations to a reviewer’s own articles are justified as they may highlight important errors or missing context in the article. Citations to a reviewer’s own articles can be justified when the authors have made a “large scholarly oversight” [8]. To investigate this, we compared the recommendations and wording of reviewers who included citations to their own articles to reviewers who included citations to other articles. The language used by the reviewers of these two groups was similar, with a higher use of “please” and “need” when not approving the article (Figure 5). However, there was a difference between groups in their recommendations, as reviewers including citations to other articles were more likely to recommend “Not approved” (Figure 4) whereas this association was not observed for reviewers including citations to their own articles (Figure 3). This indicates that missing citations to other articles were considered more serious than missing citations to the reviewer’s articles. Reviewers who cited their own articles may have been more inclined to give authors a chance to update their article and thus potentially include the “missing” citation(s).

Examining the context of the citations to reviewers’ own articles, we found vague or non-existent justifications (Supplement S.7), showing that some reviewers ignored the journals’ guidelines to state their reasoning when including citations to their own articles. However, these examples of poor justifications do not mean that all self-citations are coercive.

For both research questions, the effects were stronger for the second and later versions of the article than for the first version. Reviewers may understand that authors may be more willing to compromise on later versions when they are closer to obtaining an indexed publication. Most researchers understand that the peer review system is imperfect and that they sometimes have to make compromises to be successful [20, 45]. Another difference to consider is that later versions will include more articles with disagreements between reviewers and more that were not “Approved” in the first version as articles where two or more reviewers recommended “Approved” may not have needed a second version.

Potential improvements to peer review

Journals could give stronger guidance to reviewers and authors on coercive citations [4]. However, given the limited time for peer review and the many differences in guidelines between journals [46], most authors may not read peer review instructions. Hence, guidance alone may have limited impact.

One suggestion is that reviewers declare to editors when they have recommended citations to their own work [47]. A useful innovation would be for all reviews that contain citations to the reviewer’s own articles to be automatically flagged to the editors who could check if the citations are justified. We are aware of one journal where this is already happening (personal communication, Benno Torgler). F1000 have recently introduced checks to prevent reviewers from publishing a review with three or more citations to the reviewer’s own articles. If the reviewers continue to request more than three, then the review is examined, and if the citations are deemed inappropriate and the reviewer declines to remove them, then the review is declined.

Open peer review has been suggested as a way to reduce coercive citations [7, 47]. However, our results from four journals that use open review show that it is not a perfect antidote, although the problem could be worse in journals using blinded peer review. The transparency of open peer review should prevent reviewers from leaving self-serving comments; however, we found some dubious justifications for self-citations and blatant use of AI (Supplement S.7). These reviewers may have rationalised that although their words are public, they are rarely scrutinised (Supplement S.8); hence, it was worth the risk. The assumed additional quality assurance from open peer review [48] may often be absent.

A more radical change to peer review is that the reviewers initially see a version of the article with all references blinded and no reference list; for example, “A strong association between increased cleaning and reduced hospital infection is well established [x]”. Reviewers are asked to give an initial recommendation and comments, and then are shown the version with the full references and asked if they need to update their recommendation or provide additional comments. However, this involves more administrative work and demands more from peer reviewers. This approach could be used for particularly consequential or controversial articles. Some journals already require authors to partially blind their articles to maintain anonymous peer review; for example, the instructions from Taylor & Francis include blinding the authors’ names in the reference list [49].

An argument could be made for using large language models to provide peer review that is unmoved by citation flattery. However, peer review is an inherently human task by peers, and instead we need to improve peer review rather than abdicating this often difficult and time-consuming task to machines [50].

Related research

Previous cross-sectional studies of self-citations in reviews found at least one self-citation in 3% at a journal that used blinded peer review [17], 12% at a journal that used blinded peer review [51], and 12% at a journal that used open peer review [52]. A related study found that 15% of reviews included a self-citation and that the self-citations were highest when the reviewer recommended “major revisions” [53]. These figures are comparable with the 6% found here and indicate that most reviews do not include self-citations.

Previous surveys estimated that 14% and 20% of authors had experienced a coercive citation request from an editor [54, 55], and 7% and 23% had experienced coercive citation pressure from a reviewer [14, 56]. The frequency with which researchers interact with peer review means that many will encounter coercive citations at some point in their career.

A study of conference submissions estimated that reviewers who were cited gave submissions much higher scores [18]. A study of journal peer review estimated that cited reviewers scored the article higher, but with potential confounding by the quality of the article [17].

A survey of authors concluded that accepting an editor’s request for citations improved the chances of being accepted [12]. Requests in later versions were more strongly associated with acquiescence, and we found a related pattern in our analysis, with reviewers who included citations to their own articles being much less likely to recommend approval for later versions (Figure 3). A study examining open peer review found that requests to cite the reviewer’s articles were more likely to be included than other suggested citations, indicating that many authors wanted to please the reviewer or felt pressure to do so [52].

A survey of journal editors found that only 5% objected to reviewers citing their own articles, and that this should be expected as reviewers are likely to have done related work [8].

A cross-sectional study found that reviewers citations to their own articles were more likely to have no rationale compared to other citations, suggesting that they are more likely to be unwarranted [51].

Strengths and limitations

This is an observational study meaning we cannot rule out unmeasured confounding and should be cautious in interpreting the results.

To our knowledge, this is the first analysis to use a matched design and analysis when examining reviewer citations, and hence strongly control for any confounding by the characteristics of the authors or articles. We compared reviewers who examined an identical article; hence, the differences we found should be due to the reviewers.

Our models include measurement error, as some citations to the reviewers’ work will be missed by our data collection, and some captured citations will be inaccurate [57]. We performed random data checks that showed good accuracy (Supplement S.9); however, we also found valid citations that were not captured by our data extraction for conference proceedings and technical reports, which are less likely to have a DOI. This measurement error would most likely underestimate a true association, as it reduced the variance in citation counts and created a regression dilution [58]. Our estimates will be biased if the associations between citations and reviewers’ recommendations are different for publications that do not have a DOI. Reviewers should be equally happy with any citation to their work; however, some reviewers may prefer citations to indexed articles, as these are more likely to count toward their h-indices [59].

We examined whether citing a reviewer altered their recommendation, but did not examine the sentiment of the citation [60]. Some citations would likely have been critical of the reviewer’s articles, and we would expect these to reduce the chances of a favourable recommendation. An analysis that included the sentiment of the citation would be useful, although previous research found that most citations are neutral or positive [60].

We did not examine the authors’ responses to the reviewers but these could include important information on why a citation was included or not in a revised version of the article. A detailed analysis examining the text used in the interactions between authors and reviewers could provide valuable information about the peer review process.

Our results may not be generalisable to journals that use blinded peer review or journals that use the traditional peer review model rather than the publish–review–curate model studied here. A previous study found that asking reviewers to consent to an open review had no important effect on the quality of the review or the reviewers’ recommendation [61]. Another potential difference is that the journals in our sample often asked the authors to suggest peer reviewers; however, this is relatively common in other journals [8].

We found a bias in our sample, as co-reviewers and reviewers from older articles were more likely to be excluded due to not having an OpenAlex record (Supplement S.6). We therefore lost more junior reviewers who were less likely to be cited. The percentage of reviews lost was 5% (2,026 of 39,113), which is hopefully small enough to avoid a large bias.

eLife Assessment

This important study explored a number of issues related to citations in the peer review process. An analysis of more than 37000 peer reviews at four journals found that: i) during the first round of review, reviewers were less likely to recommend acceptance if the article under review cited the reviewer's own articles; ii) during the second and subsequent rounds of review, reviewers were more likely to recommend acceptance if the article cited the reviewer's own articles; iii) during all rounds of review, reviewers who asked authors to cite the reviewer's own articles (a practice known as 'coercive citation') were less likely to recommend acceptance. However, when an author agreed to cite work by the reviewer, the reviewer was more likely to recommend acceptance of the revised article. The evidence is convincing, and while the revisions made by the author have addressed most of the concerns the reviewers had about the original version, a small number of concerns remain.

Reviewer #1 (Public review):

Summary:

The work used open peer reviews and followed them through a succession of reviews and author revisions. It assessed whether a reviewer had requested the author include additional citations and references to the reviewers' work. It then assessed whether the author had followed these suggestions and what the probability of acceptance was based on the authors decision. Reviewers who were cited were more likely to recommend the article for publication when compared with reviewers that were not cited. Reviewers who requested and received a citation were much likely to accept than reviewers that requested and did not receive a citation.

Strengths and weaknesses:

The work's strengths are the in-depth and thorough statistical analysis it contains and the very large dataset it uses. The methods are robust and reported in detail.

I am still concerned that there is a major confounding factor: if you ignore the reviewers requests for citations are you more likely to have ignored all their other suggestions too? This has now been mentioned briefly and slightly circuitously in the limitations section. I would still like this (I think) major limitation to be given more consideration and discussion, although I am happy that it cannot be addressed directly in the analysis.

Reviewer #2 (Public review):

Summary:

This article examines reviewer coercion in the form of requesting citations to the reviewer's own work as a possible trade for acceptance and shows that, under certain conditions, this happens.

Strengths:

The methods are well done and the results support the conclusions that some reviewers "request" self-citations and may be making acceptance decisions based on whether an author fulfills that request.

Weakness:

I thank the author for addressing my comments about the original version.

Reviewer #3 (Public review):

Summary:

In this article, Barnett examines a pressing question regarding citing behavior of authors during the peer review process. In particular, the author studies the interaction between reviewers and authors, focusing on the odds of acceptance, and how this may be affected by whether or not the authors cited the reviewers' prior work, whether the reviewer requested such citations be added, and whether the authors complied/how that affected the reviewer decision-making.

Strengths:

The author uses a clever analytical design, examining four journals that use the same open peer review system, in which the identities of the authors and reviewers are both available and linkable to structured data. Categorical information about the approval is also available as structured data. This design allows a large scale investigation of this question.

Weaknesses:

My original concerns have been largely addressed. Much more detail is provided about the number of documents under consideration for each analysis, which clarifies a great deal.

Much of the observed reviewer behavior disappears or has much lower effect sizes depending on whether "Accept with Reservations" is considered an Accept or a Reject. This is acknowledged in the results text. Language has been toned down in the revised version.

The conditional analysis on the 441 reviews (lines 224-228) does support the revised interpretation as presented.

No additional concerns are noted.

Reviewer #4 (Public review):

Summary:

This work investigates whether a citation to a referee made by a paper is associated with a more positive evaluation by that referee for that paper. It provides evidence supporting this hypothesis. The work also investigates the role of self-citations by referees where the referee would ask authors to cite the referee's paper.

Strengths:

This is an important problem: referees for scientific papers must provide their impartial opinions rooted in core scientific principles. Any undue influence due to the role of citations breaks this requirement. This work studies the possible presence and extent of this.

The methods are solid and well done. The work uses a matched pair design which controls for article-level confounding and further investigates robustness to other potential confounds.

Weaknesses:

The authors have addressed most concerns in the initial review. The only remaining concern is the asymmetric reporting and highlighting of version 1 (null result) versus version 2 (rejecting null). For example the abstract says "We find that reviewers who were cited in the article under review were more likely to recommend approval, but only after the first version (odds ratio = 1.61; adjusted 99.4% CI: 1.16 to 2.23)" instead of a symmetric sentence "We find ... in version 1 and ... in version 2"

Author response:

The following is the authors’ response to the previous reviews.

Editors comments:

I would encourage you to submit a revised version that addresses the following two points:

[a] The point from Reviewer #1 about a possible major confounding factor. The following article might be germane here: Baas and Fennell, 2019: https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3339568

I don’t believe that the point raised by reviewer 1 is a confounder, see my response below.

This article highlighted was in my reading list, but I did not cite it because I was confused by its methods.

The point from Reviewer #4 about the abstract. It is important that the abstract says something about how reviewers reacted to the original versions of articles in which they were cited (ie, the odds ratio = 0.84, etc result), before going on to discuss how they reacted to revised articles (ie, the odds ratio = 1.61, etc result). I would suggest doing this along the following lines - but please feel free to reword the passage "but this effect was not strong/conclusive":

When reviewers were cited in the original version of the article under review, they were less likely to approve the article compared with reviewers who were not cited, but this effect was not strong/conclusive (odds ratio = 0.84; adjusted 99.4% CI: 0.69-1.03). However, when reviewers were cited in the revised version of the article, they were more likely to approve compared with reviewers who were not cited (odds ratio = 1.61; adjusted 99.4% CI: 1.16-2.23).

I have changed the abstract to include the odds ratios for version 1 and have used the same wording as from the main text.

Reviewer #1 (Public review):

Summary:

The work used open peer reviews and followed them through a succession of reviews and author revisions. It assessed whether a reviewer had requested the author include additional citations and references to the reviewers' work. It then assessed whether the author had followed these suggestions and what the probability of acceptance was based on the authors decision. Reviewers who were cited were more likely to recommend the article for publication when compared with reviewers that were not cited. Reviewers who requested and received a citation were much likely to accept than reviewers that requested and did not receive a citation.

Strengths and weaknesses:

The work's strengths are the in-depth and thorough statistical analysis it contains and the very large dataset it uses. The methods are robust and reported in detail.

I am still concerned that there is a major confounding factor: if you ignore the reviewers requests for citations are you more likely to have ignored all their other suggestions too? This has now been mentioned briefly and slightly circuitously in the limitations section. I would still like this (I think) major limitation to be given more consideration and discussion, although I am happy that it cannot be addressed directly in the analysis.

This is likely to happen, but I do not think it’s a confounder. A confounder needs to be associated with both the outcome and the exposure of interest. If we consider forthright authors who are more likely to rebuff all suggestions, then they would receive just as many citation and self-citation requests as authors who were more compliant. The behaviour of forthright authors would likely only reduce the association seen in most authors which would be reflected in the odds ratios.

Reviewer #2 (Public review):

Summary:

This article examines reviewer coercion in the form of requesting citations to the reviewer's own work as a possible trade for acceptance and shows that, under certain conditions, this happens.

Strengths:

The methods are well done and the results support the conclusions that some reviewers "request" self-citations and may be making acceptance decisions based on whether an author fulfills that request.

Weakness:

I thank the author for addressing my comments about the original version.

Reviewer #3 (Public review):

Summary:

In this article, Barnett examines a pressing question regarding citing behavior of authors during the peer review process. In particular, the author studies the interaction between reviewers and authors, focusing on the odds of acceptance, and how this may be affected by whether or not the authors cited the reviewers' prior work, whether the reviewer requested such citations be added, and whether the authors complied/how that affected the reviewer decision-making.

Strengths:

The author uses a clever analytical design, examining four journals that use the same open peer review system, in which the identities of the authors and reviewers are both available and linkable to structured data. Categorical information about the approval is also available as structured data. This design allows a large scale investigation of this question.

Weaknesses:

My original concerns have been largely addressed. Much more detail is provided about the number of documents under consideration for each analysis, which clarifies a great deal.

Much of the observed reviewer behavior disappears or has much lower effect sizes depending on whether "Accept with Reservations" is considered an Accept or a Reject. This is acknowledged in the results text. Language has been toned down in the revised version.

The conditional analysis on the 441 reviews (lines 224-228) does support the revised interpretation as presented.

No additional concerns are noted.

Reviewer #4 (Public review):

Summary:

This work investigates whether a citation to a referee made by a paper is associated with a more positive evaluation by that referee for that paper. It provides evidence supporting this hypothesis. The work also investigates the role of self-citations by referees where the referee would ask authors to cite the referee's paper.

Strengths:

This is an important problem: referees for scientific papers must provide their impartial opinions rooted in core scientific principles. Any undue influence due to the role of citations breaks this requirement. This work studies the possible presence and extent of this.

The methods are solid and well done. The work uses a matched pair design which controls for article-level confounding and further investigates robustness to other potential confounds.

Weaknesses:

The authors have addressed most concerns in the initial review. The only remaining concern is the asymmetric reporting and highlighting of version 1 (null result) versus version 2 (rejecting null). For example the abstract says "We find that reviewers who were cited in the article under review were more likely to recommend approval, but only after the first version (odds ratio = 1.61; adjusted 99.4% CI: 1.16 to 2.23)" instead of a symmetric sentence "We find ... in version 1 and ... in version 2".

The latest version now includes the results for both versions.

Version 3

Abstract

Peer reviewers sometimes comment that their own journal articles should be cited by the journal article under review. Comments concerning relevant articles can be justified, but comments can also be unrelated coercive citations. Here, we used a matched observational study design to explore how citations influence the peer review process. We used a sample of more than 37,000 peer reviews from four journals that use open peer review and make all article versions available. We find that reviewers who were cited in versions after version 1 were more likely to make a favourable recommendation (odds ratio = 1.61; adjusted 99.4% CI: 1.16 to 2.23), whereas being cited in the first version did not improve their recommendation (odds ratio = 0.84; adjusted 99.4% CI: 0.69 to 1.03). For all versions of the articles, the reviewers who commented that their own articles should be cited were less likely to recommend approval compared to the reviewers who did not, with the strongest association after the first version (odds ratio = 0.15; adjusted 99.4% CI: 0.08 to 0.30). Reviewers who included a citation to their own articles were much more likely to approve a revised article that cited their articles compared to a revised article that did not (odds ratio = 3.5; 95% CI: 2.0 to 6.1). Some reviewers’ recommendations depend on whether they are cited or want to be cited. Reviewer citation requests can turn peer review into a transaction rather than an objective critique of the article.

Introduction

In 2024, a published peer reviewed article included this remarkable sentence: “As strongly requested by the reviewers, here we cite some references [35–47] although they are completely irrelevant to the present work” [1]. This was a rare public example of coerced citations, where a reviewer exploits the peer review process to increase their citation counts and hence further their own career [2–4]. Reviewers should be relevant experts, so some suggestions to cite their articles will be appropriate. However, excessive citation requests or requests to cite unrelated articles are unethical [5–9]. Coerced citations can also come from editors trying to boost their journal’s ranking [10–12].

Coerced citations are reported as a common problem in peer review. In author surveys, two-thirds reported pressure from peer reviewers to cite unrelated articles [13] and 23% had experienced a reviewer that “required them to include unnecessary references to their publication(s)” [14]. Publishers have investigated whether “hundreds of researchers” have manipulated the peer review process to increase their own citations [15]. Some reviewers may be exploiting their power over authors who “have a strong incentive to […] accept all ‘suggestions’ by the referees even if one knows that they are misleading or even incorrect” [16].

As reviewers are often in the same field as the article’s authors, they may already be cited in the article without the need for coerced citations. Reviewers who are cited may give a more favourable peer review and be more willing to overlook flaws [17, 18]. Some authors may try to exploit this using “referee baiting” [3] or “flattery citations” [19] by favourably citing a reviewer’s work.

The interactions during peer review between authors and reviewers can determine whether an article is accepted [20] and what results are included in the published version [21]. Given the importance of peer review for science, studies that examine how peer review works in practice are needed [22–26]. Here, we examine interactions between peer reviewers and authors using four journals that publish all article versions and all peer reviews. We had two research questions:

Do peer reviewers give a more or less favourable recommendation when they are cited in the article?

Do peer reviewers give a more or less favourable recommendation when their review includes a citation to their own articles?

Methods

Journal selection

We studied journals from the publisher F1000 as their journals use open peer review with signed reviewers. F1000 journals use a publish-review-curate model [27], meaning all versions of the article are publicly available, including versions updated after peer review. This allowed us to examine the interactions between authors and reviewers throughout the peer review process. We selected four F1000 journals that each had over 100 articles. Some characteristics of the four journals are given in Table 1. Three journals were created to support funders.

The peer review process used by F1000 journals differs from most standard journals. The journals do not use academic editors, but do have in-house editors who manage articles but do not make editorial decisions. This means that most interactions during peer review are between authors and reviewers directly. In-house editors perform checks prior to the first version of the article being published and at F1000Research this results in 40 to 50% submissions being rejected (personal communication, F1000 staff). Up to mid-2024, authors were asked to identify potential reviewers who were qualified experts with no competing interests [28]. Since mid-2024, reviewer identification is made in-house, although authors can suggest reviewers.

Reviewers are asked to recommend one of three categories: Approved, Approved with reservations, and Not approved. For brevity, we refer to ‘Approved with Reservations’ as ‘Reservations’. An article is indexed once it receives two ‘Approved’ or two ‘Reservations’ and one ‘Approved’. The guidelines for recommending Approved are: “the aims and research methods are adequate; results are presented accurately, and the conclusions are justified and supported by the presented data” [29]. Peer reviewers are asked to assess the validity of an article’s content, rather than novelty or interest levels, an approach designed to combat publication bias [30].

All four journals have a peer reviewer code of conduct and state that reviewers should familiarise themselves with the ethical guidelines for peer reviewers by the Committee On Publication Ethics [31]. The journals’ guidelines for reviewers include the following: “reviewers should explicitly state their reasoning when asking authors to cite their own work”.

Data extraction

We extracted data on authors and articles from the OpenAlex database (https://openalex.org/) and directly from the four journals. OpenAlex combines scholarly data from multiple sources, including ORCID - a unique identifier for researchers, Microsoft Academic, Crossref and PubMed. A recent study compared OpenAlex with the two commonly used proprietary bibliometric databases of Web of Science and Scopus for the years 2015 to 2022 [32]. The results were mixed, but OpenAlex had better ORCID coverage and covered more Digital Object Identifiers (DOIs) - the unique identifier for publications. We accessed OpenAlex using the openalexR package [33, 34]. We used each journal’s application programming interface (API) to extract data on the articles and peer reviews. The data were extracted in four stages:

Searches were made using the APIs of the four journals to find all articles published between 1 Jan 2012 and 28 May 2025, with the start date to capture all potential articles.

For each article, the following data were downloaded in XML format:

The article’s publication date and version number

The reviewers’ names and ORCIDs (if available)

The text of all reviews and the reviewers’ recommendations

The DOIs and PMIDs (PubMed IDs) from the article’s reference list

The DOIs and PMIDs of any articles cited by the reviewers. The online peer review system at F1000 journals includes the DOI of any article cited in the review, which facilitates the identification of citations to the reviewers’ articles.

Articles were excluded if:

They were not peer reviewed or had yet to receive any reviews

The reference list was empty

The reviewers’ publication histories were collected from OpenAlex using their name, institution and ORCID (if available). Reviews were excluded if there was no record for the reviewer in OpenAlex, or if the reviewer had no published articles as there was no potential for them to be cited or request a citation to their own articles.

Study design

We used two predictor variables about the reviewer:

The number of times they were cited in the article (0,1, 2,…).

The number of times they included citations to their own articles in their review (0,1, 2,…).

We fitted both predictors as linear, but reviewers may behave differently with any citation rather than a linear change, and hence we also fitted both predictors as a binary “none versus any” (0 versus 1, 2,…). We compared the linear and binary alternatives using the Akaike Information Criterion (AIC) to find the parameterisation that best fitted the data [35].

We matched on article and version to control for confounding by any characteristics of the article [36]; for example, the article’s topic or writing style. Hence, we compared two or more independent reviewers who considered the same article.

All analyses were stratified by article version, using the first version only or the second and subsequent versions. This is because the reviewers are unknown to the authors for the first version, but from the second version onwards, the authors will know the reviewers as the journals use signed peer reviews. This knowledge could alter the behaviour of authors and reviewers.

The study design is summarised in Figure 1.

Graphical summary of the study design for research question 1 showing a dummy article and two reviews.

In the first version of the article, the reviewer Smith (blue) is cited whilst Jones (purple) is not. For the second version of the article, the authors are now aware that Jones is a reviewer and Jones has been cited. The reviewers’ recommendations are the outcome and are colour-coded as Not approved (red), Reservations (orange) and Approved (green). We tested whether citations to the reviewer in the article influenced their recommendation. The matched design means that only reviewers of the same article are compared (here, Smith and Jones) and the overall effect is estimated by aggregating over multiple matched comparisons. Research question 2 used the same design but examined citations to the reviewers’ articles in their reviews.

Statistical methods

We used conditional logistic regression to examine the associations between the citations to the reviewer and their ordinal recommendation (Approved → Reservations → Not approved) while matching the article and the version [37]. Conditional logistic regression requires a binary dependent variable; hence, we fitted two related models that examined the odds of:

“Approved” compared with “Reservations” or “Not approved”.

“Approved” or “Reservations” compared with “Not approved”,

These two models tested the same hypothesis, hence we adjusted for multiple testing. We also used repeated testing due to the stratification by article version and the two formulations of the predictors (linear or none versus any). Since we used 8 (2 × 2 × 2) tests, we displayed all the results using 99.4% confidence intervals instead of 95.0% intervals, which is a 5% type I error divided by eight tests.

In an unplanned analysis, we examined the association between the reviewer’s recommendation and whether they included citations to work other than citations to their own articles. This was added to examine differences between reviewers’ citations to their own articles and other articles.

Outliers were not excluded. No data were missing in the analysis data set.

The sample size calculation is in Supplement S.1.

Text analysis

We examined how reviewers’ citations to their own articles or other articles were justified and whether their wording differed according to their recommendation. For an initial view of citations to their own articles, we randomly selected 20 reviews and extracted the most relevant sentence concerning the citation.

To analyse the review text, we first extracted the 100 most commonly used words in all reviews. To standardise the text, all words were transformed into tokens, with stop-words removed and then stemmed. We then tested which of the 100 words were associated with recommending Approved versus Reservations or Not approved amongst those reviewers who included a citation to their own articles and those who included a citation to other articles. We chose the set of words using an elastic net with 10-fold cross-validation and selected a parsimonious model by using the lambda within one standard error of the minimum cross-validated error [38, 39]. To get uncertainty intervals for the estimates, we fitted a Bayesian model with the set of words selected by the elastic net and using a sceptical Normal prior centred on zero to create shrinkage.

Reproducibility

Research question 1 was pre-registered using As Predicted on 20 May 2024 [40]. Research question 2 was formulated during data collection but before any data analysis and used the same study design and statistical methods as question 1.

All data extraction and analyses were conducted using R version 4.4.1 [41]. The data and R code are available on GitHub [42].

Results

A flow chart of the included reviews is shown in Supplement S.2. The final sample size was over 37,000 reviews. There were more than 3,500 articles that were not included because they had not yet been peer reviewed, especially recent articles. More than 2,000 reviewers did not have a record in OpenAlex and so could not be included. These missing reviewers were more likely to be from older articles and more likely to be co-reviewers.

Descriptive statistics on the included reviews are in Table 2. The reviewers were cited at least once in 13% of the articles and 6% of the reviews included a self-citation. Most reviews recommended “Approved” (54%), with only 8% recommending “Not approved” which is low compared with many journals; however, 40 to 50% of submissions are rejected before articles are sent for peer review (personal communication, F1000 staff). The reviewers were relatively experienced, with a median number of papers of 55.

The binary predictor for citations of “any versus none” had a generally better fit to the data compared to the linear predictor (Supplement S.3). This indicates that for most reviewers receiving any citation is important, and there is no linear increase for two or more citations. The following results are for the binary predictor “any versus none”, with the results using a linear predictor in Supplement S.4.

Reviewers who were cited were more likely to approve the article, but only after version 1 (Fig 2 and Table 3). If a reviewer was cited in any versions after version 1, the odds ratio for recommending Approved versus Reservations or Not approved was 1.61 (adjusted 99.4% CI 1.16 to 2.23).

Odds ratios and probabilities for reviewers giving a more or less favourable recommendation depending on whether they were cited in the article.

Top left: Odds ratios for reviewers giving a more favourable (Approved) or less favourable (Reservations or Not approved) recommendation depending on whether they were cited in the article. Reviewers cited in later versions (blue) were more likely to make a favourable recommendation (odds ratio = 1.61; adjusted 99.4% CI: 1.16 to 2.23), whereas being cited in the first version (green) did not improve their recommendation (odds ratio = 0.84; adjusted 99.4% CI: 0.69 to 1.03). Top right: Same results as top left displayed as conditional probabilities. From the top, the lines show the within-strata probability of a reviewer approving: a version 1 article in which they are not cited (0.51; adjusted 99.4% CI: 0.49 to 0.52); a version 1 article in which they are cited (0.46; adjusted 99.4% CI: 0.41 to 0.51); a version 2 (or higher) article in which they are not cited (0.48; adjusted 99.4% CI: 0.45 to 0.51); and a version 2 (or higher) article in which they are cited (0.60; adjusted 99.4% CI: 0.53 to 0.65). Bottom left: Same estimates as top left except that a more favourable recommendation is now Approved or Reservations and a less favourable is Not approved. There was no clear association for cited reviewers in version 1 (odds ratio = 0.84; adjusted 99.4% CI: 0.57 to 1.23) or later versions (odds ratio = 1.12; adjusted 99.4% CI: 0.59 to 2.13). Bottom right: Same results as bottom left displayed as conditional probabilities. From the top, the lines show the within-strata probability of a reviewer approving: a version 1 article in which they are not cited (0.51; adjusted 99.4% CI: 0.48 to 0.53); a version 1 article in which they are cited (0.47; adjusted 99.4% CI: 0.35 to 0.55); a version 2 (or higher) article in which they are not cited (0.50; adjusted 99.4% CI: 0.43 to 0.55); and a version 2 (or higher) article in which they are cited (0.53; adjusted 99.4% CI: 0.34 to 0.63). This figure is based on an analysis of 12,051 articles and 24,677 reviews for version 1 and 6090 articles and 10,196 reviews for version 2+. In all panels a dot or square represents a mean, and a horizontal line represents an adjusted 99.4% confidence interval.

Reviewers who included a citation to their own articles were much less likely to approve the article for all versions (Fig 3 and Table 3). The odds ratio for recommending Approved versus Reservations or Not approved was 0.57 (99.4% CI 0.44 to 0.73) for version 1 and strengthened to 0.15 (99.4% CI 0.08 to 0.30) for versions 2+. The less favourable recommendation was only for the approval of the article and the odds ratios for Approved or Reservations versus Not approved were much closer to 1.

Odds ratios and probabilities for reviewers giving a more or less favourable recommendation if they included a citation to their own articles in their review.

Top left: Odds ratios for reviewers giving a more favourable (Approved) or less favourable (Reservations or Not approved) recommendation depending on whether their review included a citation to their own articles. Reviewers including a citation to their own articles were less likely to make a favourable recommendation for version 1 (green; odds ratio = 0.57; adjusted 99.4% CI: 0.44 to 0.73) and later versions (blue; odds ratio = 0.15; adjusted 99.4% CI: 0.07 to 0.30). Top right: Same results as top left displayed as conditional probabilities. From the top, the lines show the within-strata probability of a reviewer approving: a version 1 article in which their review did not include a citation (0.51; adjusted 99.4% CI: 0.50 to 0.53); a version 1 article in which in which their review included a citation (0.37; adjusted 99.4% CI: 0.29 to 0.44); a version 2 (or higher) article in which their review did not include a citation (0.51; adjusted 99.4% CI: 0.49 to 0.53); and a version 2 (or higher) article in which in which their review included a citation (0.14; adjusted 99.4% CI: 0.01 to 0.30). Bottom left: Same estimates as top left except that a more favourable recommendation is now Approved or Reservations and a less favourable is Not approved. There was no clear association for reviewers who included a citation to their own articles in version 1 (odds ratio = 1.11; adjusted 99.4% CI: 0.77 to 1.60) or later versions (odds ratio = 0.80; adjusted 99.4% CI: 0.37 to 1.74). Bottom right: Same results as bottom left displayed as conditional probabilities. From the top, the lines show the within-strata probability of a reviewer approving: a version 1 article in which their review did not include a citation (0.50; adjusted 99.4% CI: 0.48 to 0.52); a version 1 article in which their review included a citation (0.53; adjusted 99.4% CI: 0.43 to 0.60); a version 2 (or higher) article in which their review did not include a citation (0.51; adjusted 99.4% CI: 0.46 to 0.54); and a version 2 (or higher) article in which their included a citation (0.45; adjusted 99.4% CI: 0.12 to 0.61). This figure is based on an analysis of 12,078 articles and 24,732 reviews for version 1 and 6101 articles and 10,213 reviews for version 2+. In all panels a dot or square represents a mean, and a horizontal line represents an adjusted 99.4% confidence interval.

In an unplanned analysis, we examined the behaviour of reviewers in the first two versions of the article. We examined the 441 reviews where the reviewer was not cited in version 1 of the article and included a citation to their own articles in their first review. The reviewers who were then cited in version 2 recommended approval for 92% compared to 76% for reviewers who were not cited (odds ratio = 3.5, 95% CI: 2.0 to 6.1). This analysis did not use matching.

In an unplanned analysis, we examined whether the reviewers’ recommendations depended on whether their review included citations to articles other than their own. Reviewers who included citations in their review were much more likely not to approve the article (Figure 4), which was similar to the association with citations to reviewers’ own articles (Figure 3). However, reviewers who included citations to articles other than their own were also much more likely to recommend “Not approved”, as shown by the lower odds of “Approved” or “Reservations” versus “Not approved”. This association was not seen using citations to reviewers’ own articles (Figure 3)

Odds ratios and probabilities for reviewers giving a more or less favourable recommendation depending on if they included citations to articles other than their own in their review.

Top left: Odds ratios for reviewers giving a more favourable (Approved) or less favourable (Reservations or Not approved) recommendation depending on whether their review included a citation to articles other than their own. Reviewers including citations to other articles were less likely to make a favourable recommendation for version 1 (green; odds ratio = 0.53; adjusted 99.4% CI: 0.44 to 0.64) and later versions (blue; odds ratio = 0.18; adjusted 99.4% CI: 0.10 to 0.30). Top right: Same results as top left displayed as conditional probabilities. From the top, the lines show the within-strata probability of a reviewer approving: a version 1 article in which their review did not cite other articles (0.53; adjusted 99.4% CI: 0.51 to 0.54); a version 1 article in which their review cited other articles (0.37; adjusted 99.4% CI: 0.31 to 0.42); a version 2 (or higher) article in which their review did not cite other articles (0.52; adjusted 99.4% CI: 0.50 to 0.54); and a version 2 (or higher) article in which in which their review cited other articles (0.17; adjusted 99.4% CI: 0.02 to 0.30). Bottom left: Same estimates as top left except that a more favourable recommendation is now Approved or Reservations and a less favourable is Not approved. Reviewers including citations to other articles were less likely to make a favourable recommendation for version 1 (odds ratio = 0.62; adjusted 99.4% CI: 0.46 to 0.84) and later versions (odds ratio = 0.34; adjusted 99.4% CI: 0.16 to 0.73). Bottom right: Same results as bottom left displayed as conditional probabilities. From the top, the lines show the within-strata probability of a reviewer approving: a version 1 article in which their review did not cite other articles (0.52; adjusted 99.4% CI: 0.49 to 0.54); a version 1 article in which their review cited other articles (0.41; adjusted 99.4% CI: 0.31 to 0.48); a version 2 (or higher) article in which their review did not cite other articles (0.52; adjusted 99.4% CI: 0.47 to 0.55); and a version 2 (or higher) article in which their review cited other articles (0.27; adjusted 99.4% CI: 0.02 to 0.45). This figure is based on an analysis of 12,078 articles and 24,732 reviews for version 1 and 6101 articles and 10,213 reviews for version 2+. In all panels a dot or square represents a mean, and a horizontal line represents an adjusted 99.4% confidence interval.

Sensitivity analyses

The odds ratios when including co-reviewers with reviewers were similar to the odds ratios when using reviewers only (Supplement S.5).

We found no evidence that the reviewers’ publication numbers or country confounded the associations between citations and recommendations (Supplement S.6).

Text analyses of reviewers’ comments

A random sample of how reviewers included citations to their own articles found some vague justifications (Supplement S.7); for example, “Here are some additional publications you might consider referencing”. Other sentences adhered to the publisher’s guidelines for reviewers, as specific reasoning was provided for citations to their own articles [7]. One reviewer thanked the authors for a previous citation. Three reviews did not have a relevant sentence. One reviewer very likely used AI to write their review as it included the phrase “Certainly! Here are some potential review questions for the manuscript” [43]; this review included six self-citations with no justifications.

Reviewers who included a citation to their own articles or other articles were more likely to use the words “need” and “please” when not approving the article (Figure 5). In contrast, the words “genome” and “well” were the most strongly associated with the reviewers’ approval.

Words in the reviewers’ comments that were associated with approving the article or not for reviewers who included a citation to their own articles (n = 2,025) and reviewers who included citations to other articles (n = 4, 350).

The words were selected using an elastic net that started with the 100 most commonly used review words. The estimates from the elastic net are shown as empty circles and the mean estimates and 95% credible intervals from a Bayesian model are shown as a solid circle and horizontal line. Words are shown if the probability of a non-zero mean was over 0.95 for either reviewers who cited their own articles or reviewers who cited other articles. Four words were selected by the elastic net for the reviewers who cited other articles but not by the elastic net for reviewers who cited their own articles. The axis label shows the stemmed word and most common whole word in brackets.

To examine how often open peer reviews were viewed, we took a random sample of 200 reviews from the four journals and found that, on average, they were viewed just 1.2 times per year (Supplement S.8).

Discussion

Our results provide evidence that some reviewers have a transactional view of peer review, with their final approval dependent on citations to their work. Some reviewers may be exploiting the pressure on authors to “publish or perish”. Under this pressure, many authors may oblige and add the suggested citations, especially since adding another citation may only require a minor edit to their article [44]. Both sides gain from this transaction, as the authors get an indexed publication and the reviewer gets a citation.

A key question is whether citations to a reviewer’s own articles are justified as they may highlight important errors or missing context in the article. Citations to a reviewer’s own articles can be justified when the authors have made a “large scholarly oversight” [8]. To investigate this, we compared the recommendations and wording of reviewers who included citations to their own articles to reviewers who included citations to other articles. The language used by the reviewers of these two groups was similar, with a higher use of “please” and “need” when not approving the article (Figure 5). However, there was a difference between groups in their recommendations, as reviewers including citations to other articles were more likely to recommend “Not approved” (Figure 4) whereas this association was not observed for reviewers including citations to their own articles (Figure 3). This indicates that missing citations to other articles were considered more serious than missing citations to the reviewer’s articles. Reviewers who cited their own articles may have been more inclined to give authors a chance to update their article and thus potentially include the “missing” citation(s).

Examining the context of the citations to reviewers’ own articles, we found vague or non-existent justifications (Supplement S.7), showing that some reviewers ignored the journals’ guidelines to state their reasoning when including citations to their own articles. However, these examples of poor justifications do not mean that all self-citations are coercive.

For both research questions, the effects were stronger for the second and later versions of the article than for the first version. Reviewers may understand that authors may be more willing to compromise on later versions when they are closer to obtaining an indexed publication. Most researchers understand that the peer review system is imperfect and that they sometimes have to make compromises to be successful [20, 45]. Another difference to consider is that later versions will include more articles with disagreements between reviewers and more that were not “Approved” in the first version, as articles where two or more reviewers recommended “Approved” may not have needed a second version.

Potential improvements to peer review

Journals could give stronger guidance to reviewers and authors on coercive citations [4]. However, given the limited time for peer review and the many differences in guidelines between journals [46], most authors may not read peer review instructions. Hence, guidance alone may have limited impact.

One suggestion is that reviewers declare to editors when they have recommended citations to their own work [47]. A useful innovation would be for all reviews that contain citations to the reviewer’s own articles to be automatically flagged to the editors who could check if the citations are justified. We are aware of one journal where this is already happening (personal communication, Benno Torgler). F1000 have recently introduced checks to prevent reviewers from publishing a review with three or more citations to the reviewer’s own articles. If the reviewers continue to request more than three, then the review is examined, and if the citations are deemed inappropriate and the reviewer declines to remove them, then the review is declined.

Open peer review has been suggested as a way to reduce coercive citations [7, 47]. However, our results from four journals that use open review show that it is not a perfect antidote, although the problem could be worse in journals using blinded peer review. The transparency of open peer review should prevent reviewers from leaving self-serving comments; however, we found some dubious justifications for self-citations and blatant use of AI (Supplement S.7). These reviewers may have rationalised that although their words are public, they are rarely scrutinised (Supplement S.8); hence, it was worth the risk. The assumed additional quality assurance from open peer review [48] may often be absent.

A more radical change to peer review is that the reviewers initially see a version of the article with all references blinded and no reference list; for example, “A strong association between increased cleaning and reduced hospital infection is well established [x]”. Reviewers are asked to give an initial recommendation and comments, and then are shown the version with the full references and asked if they need to update their recommendation or provide additional comments. However, this involves more administrative work and demands more from peer reviewers. This approach could be used for particularly consequential or controversial articles. Some journals already require authors to partially blind their articles to maintain anonymous peer review; for example, the instructions from Taylor & Francis include blinding the authors’ names in the reference list [49].

An argument could be made for using large language models to provide peer review that is unmoved by citation flattery. However, peer review is an inherently human task by peers, and instead we need to improve peer review rather than abdicating this often difficult and time-consuming task to machines [50].

Related research

Previous cross-sectional studies of self-citations in reviews found at least one self-citation in 3% at a journal that used blinded peer review [17], 12% at a journal that used blinded peer review [51], and 12% at a journal that used open peer review [52]. A related study found that 15% of reviews included a self-citation and that the self-citations were highest when the reviewer recommended “major revisions” [53]. These figures are comparable with the 6% found here and indicate that most reviews do not include self-citations.

Previous surveys estimated that 14% and 20% of authors had experienced a coercive citation request from an editor [54, 55], and 7% and 23% had experienced coercive citation pressure from a reviewer [14, 56]. The frequency with which researchers interact with peer review means that many will encounter coercive citations at some point in their career.

A study of conference submissions estimated that reviewers who were cited gave submissions much higher scores [18]. A study of journal peer review estimated that cited reviewers scored the article higher, but with potential confounding by the quality of the article [17].

A survey of authors concluded that accepting an editor’s request for citations improved the chances of being accepted [12]. Requests in later versions were more strongly associated with acquiescence, and we found a related pattern in our analysis, with reviewers who included citations to their own articles being much less likely to recommend approval for later versions (Figure 3). A study examining open peer review found that requests to cite the reviewer’s articles were more likely to be included than other suggested citations, indicating that many authors wanted to please the reviewer or felt pressure to do so [52].

A survey of journal editors found that only 5% objected to reviewers citing their own articles, and that this should be expected as reviewers are likely to have done related work [8].

A cross-sectional study found that reviewers citations to their own articles were more likely to have no rationale compared to other citations, suggesting that they are more likely to be unwarranted [51].

Strengths and limitations

This is an observational study meaning we cannot rule out unmeasured confounding and should be cautious in interpreting the results.

To our knowledge, this is the first analysis to use a matched design and analysis when examining reviewer citations, and hence strongly control for any confounding by the characteristics of the authors or articles. We compared reviewers who examined an identical article; hence, the differences we found should be due to the reviewers.

Our models include measurement error, as some citations to the reviewers’ work will be missed by our data collection, and some captured citations will be inaccurate [57]. We performed random data checks that showed good accuracy (Supplement S.9); however, we also found valid citations that were not captured by our data extraction for conference proceedings and technical reports, which are less likely to have a DOI. This measurement error would most likely underestimate a true association, as it reduced the variance in citation counts and created a regression dilution [58]. Our estimates will be biased if the associations between citations and reviewers’ recommendations are different for publications that do not have a DOI. Reviewers should be equally happy with any citation to their work; however, some reviewers may prefer citations to indexed articles, as these are more likely to count toward their h-indices [59].

We examined whether citing a reviewer altered their recommendation, but did not examine the sentiment of the citation [60]. Some citations would likely have been critical of the reviewer’s articles, and we would expect these to reduce the chances of a favourable recommendation. An analysis that included the sentiment of the citation would be useful, although previous research found that most citations are neutral or positive [60].

We did not examine the authors’ responses to the reviewers but these could include important information on why a citation was included or not in a revised version of the article. A detailed analysis examining the text used in the interactions between authors and reviewers could provide valuable information about the peer review process.

Our results may not be generalisable to journals that use blinded peer review or journals that use the traditional peer review model rather than the publish-review-curate model studied here. A previous study found that asking reviewers to consent to an open review had no important effect on the quality of the review or the reviewers’ recommendation [61]. Another potential difference is that the journals in our sample often asked the authors to suggest peer reviewers; however, this is relatively common in other journals [8].

We found a bias in our sample, as co-reviewers and reviewers from older articles were more likely to be excluded due to not having an OpenAlex record (Supplement S.6). We therefore lost more junior reviewers who were less likely to be cited. The percentage of reviews lost was 5% (2,026 of 39,113), which is hopefully small enough to avoid a large bias.